---
title: "Ag employment"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
runtime: shiny
resource_files:
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.cpg
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.dbf
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.prj
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbn
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbx
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shp.xml
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
```

```{r loading jon docs and shapefiles}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", size = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.title = element_blank(),
        text = element_text(size = 15),
        plot.caption = element_text(hjust = 0.5, face = "italic"),
        legend.text = element_text(margin = margin(l = 2, r = 5)))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", size = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        text = element_text(size = 15),
        plot.caption = element_text(hjust = 0.5, face = "italic"),
        legend.text = element_text(margin = margin(l = 2, r = 5)))

theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        plot.title = element_text(hjust = 0.5),
        legend.title = element_blank(),
        text = element_text(size = 15),
        plot.caption = element_text(hjust = 0.5, face = "italic"),
        legend.text = element_text(margin = margin(l = 2, r = 5)))

counties.regions <- read_csv("Data/Join docs/county_regions.csv") %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc))

color.ruca <- c("Entirely rural" = "#5CA81F", "Town/rural mix" = "#C7EF99", "Urban/town/rural mix" = "#d8b365", "Entirely urban" = "#a6611a", "Minnesota" = "black")

color.pr <- c("Northwest" = "#810f7c","Northeast" = "#fe9929", "Central" = "#076324", "Seven County Mpls-St Paul" = "#d8b365", "Southwest" = "#1f78b4", "Southeast" = "#d7301f", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.counties <- scale_color_brewer(palette = "Dark2",
                       guide = guide_legend(ncol = 3))

mn_counties <- st_read("Data/Shapefiles/county shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE)
```

```{r prep master ag employment objects, include=FALSE, cache=TRUE}
master.ag.emp.county <- read_csv("Data/Employment/Master-agrelated-emp-county.csv")

master.ag.emp.comb.county <- master.ag.emp.county %>%
  group_by(periodyear, Name, countyfp, Dem_Desc, edr, planning.region) %>%
  summarise(ag.emp = sum(emp[naicstitle != "Total, All Industries"]),
            total.emp = sum(emp[naicstitle == "Total, All Industries"])) %>%
  ungroup()

master.ag.emp.ruca <- read_csv("Data/Employment/Master-agrelated-emp-ruca.csv") %>%
  mutate(Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

master.ag.emp.comb.ruca <- master.ag.emp.ruca %>%
  group_by(periodyear, Dem_Desc) %>%
  summarise(ag.emp = sum(emp[naicstitle != "Total, All Industries"]),
            total.emp = sum(emp[naicstitle == "Total, All Industries"])) %>%
  ungroup()

master.ag.emp.pr <- read_csv("Data/Employment/Master-agrelated-emp-pr.csv") %>%
  mutate(planning.region = str_replace(planning.region, ", MN", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"))

master.ag.emp.comb.pr <- master.ag.emp.pr %>%
  group_by(periodyear, planning.region) %>%
  summarise(ag.emp = sum(emp[naicstitle != "Total, All Industries"]),
            total.emp = sum(emp[naicstitle == "Total, All Industries"])) %>%
  ungroup()


master.ag.emp.edr <- read_csv("Data/Employment/Master-agrelated-emp-edr.csv") %>%
  mutate(planning.region = str_replace(planning.region, ", MN", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

master.ag.emp.comb.edr <- master.ag.emp.edr %>%
  group_by(periodyear, edr, planning.region) %>%
  summarise(ag.emp = sum(emp[naicstitle != "Total, All Industries"]),
            total.emp = sum(emp[naicstitle == "Total, All Industries"])) %>%
  ungroup()


```

<br>

# Summary of employment in ag-related industries

**Definition of ag-related industries**: This analysis cobbles together various NAICS codes that are directly related to "ag-related employment". This includes the following industries;

```{r ag-related industry list}
agrelated.ind.list <- read_csv("Data/Earnings/Master-ag-related-ind-wages-pr.csv") %>%
  distinct(naicstitle) %>%
  rename(`Ag-related NAICS categories` = 1)

kable(agrelated.ind.list, format = "html", align = "c") %>%
    kable_styling(bootstrap_options = "striped") %>%
    column_spec(1, width = "10em")

```

[Summary here]


<br>

# Analysis

[Description here]

## Change in ag employment vs. total employment {.tabset}

The charts below provide the change in farm employment and total employment by indexing the values to their 2001 levels. The trend to look for is how the fluctuations in farm employment “appear” in the total employment chart. If a regions employment is impacted by the farm economy, then we should see corresponding “bumps” or trends in the total employment charts that mimic the farm employment charts.

Overall, ag-related industries were definitely impacted by the crop commodity boom. There were significant increases in employment in these industries during that time and then a decline after 2012 in most regions. Interestingly, the largest increases are occurring in counties that are categorized as "town/rural mix". These counties are also not experiencing any sort of employment decline in these industries even after the downturn in crop commodities.

Not surprisingly, employment in ag-related industries are most prominent in Northwest, Central, Southwest, and Southeast Minnesota. Of course, it will be hard to discern what impacts increases and declines in ag-related industries are having on total employment, but evidence shows that there will be counties in the following EDRs that are being impacted the most;

* Northwest
  + EDR 1 - Northwest
  + EDR 4 - West Central
* Central
  + EDR 6E - East Central
* Southwest
  + EDR 6W - Upper Minnesota Valley
  + EDR 8 - Southwest
  + EDR 9 - South Central
  + EDR 10 - Southeast

The regions that may have areas most impacted by farm employment are located in the Northwest, Central, and Southwest regions of the state. Within those, the EDRs that should be looked at in more detail are;

<br>

```{r prep ag and total emp index, include=FALSE, cache=TRUE}
ag.emp.index.ruca <- master.ag.emp.comb.ruca %>%
  mutate(ag.emp.index = ag.emp / ag.emp[periodyear==2001],
         total.emp.index = total.emp / total.emp[periodyear==2001],
         ag.pct.total.emp = ag.emp / total.emp) %>%
  ungroup()
  
ag.emp.index.pr <- master.ag.emp.comb.pr %>%
  mutate(ag.emp.index = ag.emp / ag.emp[periodyear==2001],
         total.emp.index = total.emp / total.emp[periodyear==2001],
         ag.pct.total.emp = ag.emp / total.emp) %>%
  ungroup()

ag.emp.index.edr <- master.ag.emp.comb.edr %>%
  mutate(ag.emp.index = ag.emp / ag.emp[periodyear==2001],
         total.emp.index = total.emp / total.emp[periodyear==2001],
         ag.pct.total.emp = ag.emp / total.emp) %>%
  ungroup()

```

### RUCA

Overall, employment in ag-related industries grew during the crop commodity boom between 2008 and 2012 across all county groups, although never recovering to 2001 levels in entirely rural counties. The large variation from year to year for this county group might be due to low employment levels, making small changes show high percentages. Employment levels are less than 2,000 people in these industries in this county group. We will find out in the next chart below which will provide the percentage of total employment is in ag-related industries for this county group.

Other RUCA county groups have experienced modest growth in these industries, with 2018 employment numbers being between 1% and 16% of 2001 levels. The town/rural county group in particular has seen a lot of employment growth in these industries.

Total employment follows a trend which would be expected during a period of the Great Recession. Employment dropped during that time frame and then has been slowly recovering since 2012. However, the recovery has been slow in our entirely rural counties.  

In addition, none of the total employment trendlines seem to match the trendlines of ag employment, meaning there just might not be much impact here.

<br>

```{r charts ag and total emp index ruca, fig.width=8}
ag.emp.index.ruca.plot <- ggplot(ag.emp.index.ruca, aes(periodyear, ag.emp.index, color = Dem_Desc)) +
  geom_hline(yintercept = 1, color = "black") +
  geom_smooth(se = FALSE, size = 1) +
  geom_point_interactive(size = 2, aes(data_id = ag.emp.index, tooltip = paste(Dem_Desc, "\nYear: ", periodyear, "\nAg employment: ", comma(ag.emp), "\nTotal employment: ", comma(total.emp), "\nAg employment as % of total employment: ", percent(ag.pct.total.emp), "\nPercent change in ag employment since 2001: ", percent(ag.emp.index), "\nPercent change in total employment since 2001: ", percent(total.emp.index), sep = ""))) +
  labs(x="", y = "", color="", title = "Ag employment as a % of 2001 levels")+
  scale_y_continuous(labels=scales::percent)+
  coord_cartesian(ylim = seq(.55, 1.15, .1)) +
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_manual(values= color.ruca,
                     guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 10),
        plot.title = element_text(size = 10),
        legend.text = element_text(size = 10))

total.emp.index.ruca.plot <- ggplot(ag.emp.index.ruca, aes(periodyear, total.emp.index, color = Dem_Desc)) +
  geom_hline(yintercept = 1, color = "black") +
  geom_smooth(se = FALSE, size = 1) +
  geom_point_interactive(size = 2, aes(data_id = ag.emp.index, tooltip = paste(Dem_Desc, "\nYear: ", periodyear, "\nAg employment: ", comma(ag.emp), "\nTotal employment: ", comma(total.emp), "\nAg employment as % of total employment: ", percent(ag.pct.total.emp), "\nPercent change in ag employment since 2001: ", percent(ag.emp.index), "\nPercent change in total employment since 2001: ", percent(total.emp.index), sep = ""))) +
  labs(x="", y = "", color="", title = "Total employment as a % of 2001 levels")+
  scale_y_continuous(labels=scales::percent)+
  coord_cartesian(ylim = seq(.55, 1.15, .1)) +
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_manual(values= color.ruca,
                     guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 10),
        plot.title = element_text(size = 10))

ag.emp.index.ruca.legend <- get_legend(ag.emp.index.ruca.plot + theme(legend.box.margin = margin(0,0,0,0)))

ag.emp.index.ruca.legend.bottom <- plot_grid(NULL, ag.emp.index.ruca.legend, NULL, ncol = 3, rel_widths = c(0,.5,3))

girafe( ggobj =  plot_grid(ag.emp.index.ruca.plot+theme(legend.position = "none"), total.emp.index.ruca.plot + theme(legend.position = "none"), NULL, ag.emp.index.ruca.legend.bottom, ncol = 2, rel_heights = c(1,.1)), width_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```


### Planning Region

The Northeast region is a bit of an outlier here. Employment in ag-related industries is only a few hundred employees, therefore small changes have a big impact on percentages. In fact, employment in ag-related industries only make up .5% of total employment in that region. 

All other regions outside of the seven county metro are experiencing increases in these industries. In particular, the Northwest and Central regions have experience 28% and 26% employment growth in these industries since 2001, respectively.And there is little evidence that the Great Recession had any impact on these industries. In fact, they seemed to grow a bit during that time frame.

As expected, total employment follows a pattern that's been impacted by the Great Recession, with a drop between 2008 and 2012, and increases following that time. It's difficult to say if there are any similarities between these trendlines and the ag-related employment trendlines.

<br>

```{r charts ag and total emp index pr, fig.width=8}

ag.emp.index.pr.plot <- ggplot(ag.emp.index.pr, aes(periodyear, ag.emp.index, color = planning.region)) +
  geom_hline(yintercept = 1, color = "black") +
  geom_smooth(se = FALSE, size = 1) +
  geom_point_interactive(size = 2, aes(data_id = ag.emp.index, tooltip = paste(planning.region, "\nYear: ", periodyear, "\nAg employment: ", comma(ag.emp), "\nTotal employment: ", comma(total.emp), "\nAg employment as % of total employment: ", percent(ag.pct.total.emp), "\nPercent change in ag employment since 2001: ", percent(ag.emp.index), "\nPercent change in total employment since 2001: ", percent(total.emp.index), sep = ""))) +
  labs(x="", y = "", color="", title = "Ag employment as a % of 2001 levels")+
  scale_y_continuous(labels=scales::percent)+
  coord_cartesian(ylim = seq(.25, 1.35, .1)) +
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_manual(values= color.pr,
                     guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 10),
        plot.title = element_text(size = 10),
        legend.text = element_text(size = 10))


total.emp.index.pr.plot <- ggplot(ag.emp.index.pr, aes(periodyear, total.emp.index, color = planning.region)) +
  geom_hline(yintercept = 1, color = "black") +
  geom_smooth(se = FALSE, size = 1) +
  geom_point_interactive(size = 2, aes(data_id = ag.emp.index, tooltip = paste(planning.region, "\nYear: ", periodyear, "\nAg employment: ", comma(ag.emp), "\nTotal employment: ", comma(total.emp), "\nAg employment as % of total employment: ", percent(ag.pct.total.emp), "\nPercent change in ag employment since 2001: ", percent(ag.emp.index), "\nPercent change in total employment since 2001: ", percent(total.emp.index), sep = ""))) +
  labs(x="", y = "", color="", title = "Total employment as a % of 2001 levels")+
  scale_y_continuous(labels=scales::percent)+
  coord_cartesian(ylim = seq(.25, 1.35, .1)) +
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_manual(values= color.pr,
                     guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 10),
        plot.title = element_text(size = 10))

ag.emp.index.pr.legend <- get_legend(ag.emp.index.pr.plot + theme(legend.box.margin = margin(0,0,0,0)))

ag.emp.index.pr.legend.bottom <- plot_grid(NULL, ag.emp.index.pr.legend, NULL, ncol = 3, rel_widths = c(0,.5,3))

girafe( ggobj =  plot_grid(ag.emp.index.pr.plot+theme(legend.position = "none"), total.emp.index.pr.plot + theme(legend.position = "none"), NULL, ag.emp.index.pr.legend.bottom, ncol = 2, rel_heights = c(1,.1)), width_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```


### EDR

In the northwest region, EDR 4 - West Central looks to have their total employment impacted by ag-related industries. They have experienced significant growth, 40% higher since 2001, while also experiencing significant growth in their overall employment, 15% since 2001. They also seem to have weathered the Great Recession better than others, perhaps because of the tremendous growth in ag-related industries. As of 2018, employment in ag-related industries represented 7.6% of total employment.

EDR 1 - Northwest, is also an industry that may have experienced some impacts from ag-related industries. These industries represent 7% of their total employment, has experienced 15% growth in these industries, while their total employment has experienced a slight recovery during time of growth in ag-related industries (2012 - 2018).

EDR 5 - North Central has similar trends to EDR 4 and 1, however employment in ag-related industries is less than 5% of the total employment in the region. 

Employment in ag-related industries in EDR 2 - Headwaters only represents 3% or less of total employment and there seems to be little relation.

In the Central region, EDR 6E - East Central has 10% or more of their total employment represented in ag-related industries, has experienced employment growth in these industries, and their total employment seems to have a similar trendline.

Any of the EDRs in Southwest Minnesota could be impacted by ag-related industries. After peaking at 12%, employment in ag-related industries in EDR 6W - Upper Minnesota Valley has decreased significantly and now only represents 7% of total employment. In addition, total employment in that region is at 95% of their total employment in 2001. These trendlines are very similar.

EDR 8 - Southwest peaked at 16% of their total employment in ag-related industries and by 2018 dropped to 14%. It's a little hard to tell whether the total employment trendline is similar, but the fact that they haven't seen any growth is evidence that there is a tie here.

Lastly, Southeast Minnesota has experienced modest growth in their employment in ag-related industries as well as their total employment. By 2018, ag-related industries represented 6% of total employment.

<br>

```{r charts ag and total emp index edr, fig.height=10}

ag.emp.index.edr.plot <- ggplot(ag.emp.index.edr, aes(periodyear, ag.emp.index, color = edr)) +
  geom_hline(yintercept = 1, color = "black") +
  facet_wrap(~planning.region, ncol = 1) +
  geom_smooth(se = FALSE, size = 1) +
  geom_point_interactive(size = 2, aes(data_id = ag.emp.index, tooltip = paste(edr, "\nYear: ", periodyear, "\nAg employment: ", comma(ag.emp), "\nTotal employment: ", comma(total.emp), "\nAg employment as % of total employment: ", percent(ag.pct.total.emp), "\nPercent change in ag employment since 2001: ", percent(ag.emp.index), "\nPercent change in total employment since 2001: ", percent(total.emp.index), sep = ""))) +
  labs(x="", y = "", color="", title = "Ag employment as a % of 2001 levels")+
  scale_y_continuous(labels=scales::percent)+
  coord_cartesian(ylim = seq(.7, 1.5, .1)) +
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_manual(values= color.edr,
                     guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 10),
        plot.title = element_text(size = 10),
        legend.text = element_text(size = 6),
        axis.text.x = element_text(size = 6))


total.emp.index.edr.plot <- ggplot(ag.emp.index.edr, aes(periodyear, total.emp.index, color = edr)) +
  geom_hline(yintercept = 1, color = "black") +
  facet_wrap(~planning.region, ncol = 1) +
  geom_smooth(se = FALSE, size = 1) +
  geom_point_interactive(size = 2, aes(data_id = ag.emp.index, tooltip = paste(edr, "\nYear: ", periodyear, "\nAg employment: ", comma(ag.emp), "\nTotal employment: ", comma(total.emp), "\nAg employment as % of total employment: ", percent(ag.pct.total.emp), "\nPercent change in ag employment since 2001: ", percent(ag.emp.index), "\nPercent change in total employment since 2001: ", percent(total.emp.index), sep = ""))) +
  labs(x="", y = "", color="", title = "Total employment as a % of 2001 levels")+
  scale_y_continuous(labels=scales::percent)+
  coord_cartesian(ylim = seq(.7, 1.5, .1)) +
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_manual(values= color.edr,
                     guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 10),
        plot.title = element_text(size = 10),
        axis.text.x = element_text(size = 6))

ag.emp.index.edr.legend <- get_legend(ag.emp.index.edr.plot + theme(legend.box.margin = margin(0,0,0,0)))

ag.emp.index.edr.legend.bottom <- plot_grid(NULL, ag.emp.index.edr.legend, NULL, ncol = 3, rel_widths = c(0,.5,3))

girafe( ggobj =  plot_grid(ag.emp.index.edr.plot+theme(legend.position = "none"), total.emp.index.edr.plot + theme(legend.position = "none"), NULL, ag.emp.index.edr.legend.bottom, ncol = 2, rel_heights = c(1,.1)), height_svg = 9) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```

<br>

## Ag employment as a % of total employment {.tabset}

The charts below provide employment in ag-related industries as a percentage of total employment each year. The trend to see here is how much that percentage has changed over time.

Overall, the percentages have been relatively consistent, with more significant growth in the town/rural mix county group. There doesn't seem to be any indication that the downturn in crop commodities is impacting these industries except maybe in Southwest Minnesota.

The EDRs with the most significant amount of their employment originating from ag-related industries are;

* Northwest
  + EDR 1 - Northwest: 6.6% with a little growth since 2008
  + EDR 4 - West Central: 7.6% with a little growth since 2008 
* Central
  + EDR 6E - Southwest Central: 12% with a little growth since 2008
* Southwest
  + EDR 6W - Upper Minnesota Valley: 7% with significant decline since 2010
  + EDR 8 - Southwest: 14% with significant decline since 2010
  + EDR 9 - South Central: 9.6% with a little growth since 2010

<br>

### RUCA

Not surprisingly, ag-related industries comprise a larger percentage of total employment the more rural an area is. Interestingly, the percentage increase for all the county groups during the Great Recession, likely because those industries were hiring or not laying off while all other industries were losing employment.

Similar to the previous charts, the town/rural mix county group continues to see high percentages of their employment originate from ag-related industries.

<br>

```{r chart ag emp as % of total emp ruca}
ag.pct.total.emp.ruca.plot <- ggplot(ag.emp.index.ruca, aes(periodyear, ag.pct.total.emp, color = Dem_Desc)) +
  geom_smooth(se = FALSE, size = 2) +
  geom_point_interactive(size = 3, aes(data_id = ag.emp.index, tooltip = paste(Dem_Desc, "\nYear: ", periodyear, "\nAg employment: ", comma(ag.emp), "\nTotal employment: ", comma(total.emp), "\nAg employment as % of total employment: ", percent(ag.pct.total.emp), "\nPercent change in ag employment since 2001: ", percent(ag.emp.index), "\nPercent change in total employment since 2001: ", percent(total.emp.index), sep = ""))) +
  labs(x="", y = "", color="", title = "Ag employment as a % of total employment")+
  scale_y_continuous(labels=scales::percent)+
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_manual(values= color.ruca,
                     guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom")

girafe(ggobj = ag.pct.total.emp.ruca.plot, width_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```

### Planning Region

Overall, employment in ag-related industries seems to be relatively stable across all regions despite the decline in crop commodities.

The Southwest region has, by far, the highest percentage of total employment originating from ag-related industries. By 2018, these industries represented 11% of total employment.

The Southeast, Northwest, and Central regions all have percentages between 5% and 6%.


<br>

```{r chart ag emp as % of total emp pr}
ag.pct.total.emp.pr.plot <- ggplot(ag.emp.index.pr, aes(periodyear, ag.pct.total.emp, color = planning.region)) +
  geom_smooth(se = FALSE, size = 2) +
  geom_point_interactive(size = 3, aes(data_id = ag.emp.index, tooltip = paste(planning.region, "\nYear: ", periodyear, "\nAg employment: ", comma(ag.emp), "\nTotal employment: ", comma(total.emp), "\nAg employment as % of total employment: ", percent(ag.pct.total.emp), "\nPercent change in ag employment since 2001: ", percent(ag.emp.index), "\nPercent change in total employment since 2001: ", percent(total.emp.index), sep = ""))) +
  labs(x="", y = "", color="", title = "Ag employment as a % of total employment")+
  scale_y_continuous(labels=scales::percent)+
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_manual(values= color.pr,
                     guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom")

girafe(ggobj = ag.pct.total.emp.pr.plot, width_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```

### EDR

Just like the previous EDR chart, the EDRs we highlighted there are definitely the ones with the highest percentage of employment coming from ag-related industries. EDR 4 - West Central and EDR 1 - Northwest each have 6% to 7% of their employment coming from ag-related industries with continued growth.

EDR 6E - South Central has 12% of their employment coming from ag-related industries and has stayed relatively stable since 2001.

EDR 8 - Southwest and EDR 6W - Upper Minnesota Valley have decreasing percentage of their total employment coming from ag-related industries. EDR 8 peaked in 2010 with 16.5% of their employment in these industries while EDR 6W peaked that same year with 12%. EDR 9 - South Central has experienced slow, steady growth.

The Southeast has been relatively constant at 6%.

<br>

```{r chart ag emp as % of total emp edr, fig.height=10}

ag.pct.total.emp.edr.plot <- ggplot(ag.emp.index.edr, aes(periodyear, ag.pct.total.emp, color = edr)) +
  facet_wrap(~planning.region, ncol = 1) +
  geom_smooth(se = FALSE, size = 2) +
  geom_point_interactive(size = 3, aes(data_id = ag.emp.index, tooltip = paste(edr, "\nYear: ", periodyear, "\nAg employment: ", comma(ag.emp), "\nTotal employment: ", comma(total.emp), "\nAg employment as % of total employment: ", percent(ag.pct.total.emp), "\nPercent change in ag employment since 2001: ", percent(ag.emp.index), "\nPercent change in total employment since 2001: ", percent(total.emp.index), sep = ""))) +
  labs(x="", y = "", color="", title = "Ag employment as a % of total employment")+
  scale_y_continuous(labels=scales::percent)+
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_manual(values= color.edr,
                     guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 10))

girafe(ggobj = ag.pct.total.emp.edr.plot, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```

