---
title: "Ag wages per employment"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
runtime: shiny
resource_files:
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.cpg
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.dbf
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.prj
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbn
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbx
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shp.xml
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shx
---

<link rel="stylesheet" href="styles.css" type="text/css">


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
```

```{r loading jon docs and shapefiles}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", size = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.title = element_blank(),
        text = element_text(size = 15),
        plot.caption = element_text(hjust = 0.5, face = "italic"),
        legend.text = element_text(margin = margin(l = 2, r = 5)))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", size = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        text = element_text(size = 15),
        plot.caption = element_text(hjust = 0.5, face = "italic"),
        legend.text = element_text(margin = margin(l = 2, r = 5)))

theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        plot.title = element_text(hjust = 0.5),
        legend.title = element_blank(),
        text = element_text(size = 15),
        plot.caption = element_text(hjust = 0.5, face = "italic"),
        legend.text = element_text(margin = margin(l = 2, r = 5)))

counties.regions <- read_csv("Data/Join docs/county_regions.csv") %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         edr = str_replace(edr, "  ", " "))

edr.pr <- counties.regions %>%
  distinct(edr, .keep_all = TRUE) %>%
  select(5,6) %>%
  mutate(edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""))


color.ruca <- c("Entirely rural" = "#5CA81F", "Town/rural mix" = "#C7EF99", "Urban/town/rural mix" = "#d8b365", "Entirely urban" = "#a6611a", "Minnesota" = "black")

color.pr <- c("Northwest" = "#810f7c","Northeast" = "#fe9929", "Central" = "#076324", "Seven County Mpls-St Paul" = "#d8b365", "Southwest" = "#1f78b4", "Southeast" = "#d7301f", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.label.edr <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.counties <- scale_color_brewer(palette = "Dark2",
                       guide = guide_legend(ncol = 3))

color.ag.ind <- c("Crop Production" = "#a6cee3", "Animal Production and Aquaculture" = "#1f78b4", "Support Activities for Crop Production" = "#b2df8a", "Support Activities for Animal Production" = "#33a02c", "Food Manufacturing" = "#fb9a99", "Agricultural Implement Manufacturing" = "#e31a1c", "Farm Product Raw Material Merchant Wholesalers" = "#fdbf6f", "Farm Supplies Merchant Wholesalers" = "#ff7f00", "Pesticide, Fertilizer, and Other Agricultural Chemical Manufacturing" = "#cab2d6", "Total, All Industries" = "black", "NA" = "white")

mn_counties <- st_read("Data/Shapefiles/county shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE)
```

```{r prep master wage file, include = FALSE, cache = TRUE}
ag.ind.wages.county <- read_csv("Data/Earnings/Master-ag-related-ind-wages-counties.csv")

ag.ind.wages.pr <- read_csv("Data/Earnings/Master-ag-related-ind-wages-pr.csv") %>%
  mutate(planning.region = str_replace(planning.region, " Minnesota", ""))

ag.ind.wages.edr <- read_csv("Data/Earnings/Master-ag-related-ind-wages-edr.csv") %>%
  left_join(edr.pr, by = "edr") %>%
  mutate(planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"))

```

```{r prep master ag employment objects, include=FALSE, cache=TRUE}
master.ag.emp.county <- read_csv("Data/Employment/Master-agrelated-emp-county.csv")

master.ag.emp.comb.county <- master.ag.emp.county %>%
  group_by(periodyear, Name, countyfp, Dem_Desc, edr, planning.region) %>%
  summarise(ag.emp = sum(emp[naicstitle != "Total, All Industries"]),
            total.emp = sum(emp[naicstitle == "Total, All Industries"])) %>%
  ungroup()

master.ag.emp.ruca <- read_csv("Data/Employment/Master-agrelated-emp-ruca.csv") %>%
  mutate(Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

master.ag.emp.comb.ruca <- master.ag.emp.ruca %>%
  group_by(periodyear, Dem_Desc) %>%
  summarise(ag.emp = sum(emp[naicstitle != "Total, All Industries"]),
            total.emp = sum(emp[naicstitle == "Total, All Industries"])) %>%
  ungroup()

master.ag.emp.pr <- read_csv("Data/Employment/Master-agrelated-emp-pr.csv") %>%
  mutate(planning.region = str_replace(planning.region, ", MN", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"))

master.ag.emp.comb.pr <- master.ag.emp.pr %>%
  group_by(periodyear, planning.region) %>%
  summarise(ag.emp = sum(emp[naicstitle != "Total, All Industries"]),
            total.emp = sum(emp[naicstitle == "Total, All Industries"])) %>%
  ungroup()


master.ag.emp.edr <- read_csv("Data/Employment/Master-agrelated-emp-edr.csv") %>%
  mutate(planning.region = str_replace(planning.region, ", MN", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

master.ag.emp.comb.edr <- master.ag.emp.edr %>%
  group_by(periodyear, edr, planning.region) %>%
  summarise(ag.emp = sum(emp[naicstitle != "Total, All Industries"]),
            total.emp = sum(emp[naicstitle == "Total, All Industries"])) %>%
  ungroup()

```

```{r master combined wages and employment, include=FALSE, cache=TRUE}
ag.wages.emp.county <- ag.ind.wages.county %>%
  left_join(master.ag.emp.county[,c(1,2,5,6)], by = c("countyfp", "periodyear", "naicstitle")) %>%
  mutate(wages.emp = wages / emp)

ag.wages.emp.ruca <- ag.wages.emp.county %>%
  group_by(periodyear, Dem_Desc, naicstitle) %>%
  summarise(wages = sum(wages, na.rm = TRUE),
            emp = sum(emp, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(wages.emp = wages / emp,
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

ag.wages.emp.pr <- ag.ind.wages.pr %>%
  left_join(master.ag.emp.pr[,c(1,2,3,5)], by = c("planning.region", "periodyear", "naicstitle")) %>%
  mutate(wages.emp = wages / emp,
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"))

ag.wages.emp.edr <- ag.ind.wages.edr %>%
  left_join(master.ag.emp.edr[,c(1,2,3,5)], by = c("edr", "periodyear", "naicstitle")) %>%
  mutate(wages.emp = wages / emp,
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

```


<br>

# Summary

Overall, the highest wages per employment values are located in our most metropolitan areas where they are 32% to 45% higher than other types of areas. 

When breaking it down by region, the highest wages per employment are by far found in the seven county metro. By 2018, the wage per employment in ag-related industries was $71,678 compared to $47,067 to $51,697 in the other regions of the state. Outside of the metro, the highest wage are, not surprisingly, found in Southwest Minnesota.

When breaking it down by ag-related industry, the highest wages per employment values are found, by far, in the farm product raw material merchant wholesaler ($100,492) and farm supplies merchant wholesaler ($84,826). These two are also the fourth and fifth highest employment industries out of the 9 ag-related industries analyzed.

When analyzing where the largest share of each industry's workforce was located, our entirely urban county group was by far the largest. However, this didn't translate into the seven-county metro as having the largest share for each industry. Rather, it's metro counties in each region outside of the seven-county metro that also has a significant share of the workforce.

In terms of share of the workforce by planning region, not surprisingly the Northwest, Central and Southwest had the largest share of the workforce for each of the industries, except for the two highest paying industries - farm raw product material merchant wholesalers and farm supplies merchant wholesalers. The largest share of employment in these two industries was located in the seven county metro, where the wages per employment were also double that of the other regions.

<br>

# Definitions

**Definition of ag-related industries**: This analysis cobbles together various NAICS codes that are directly related to "ag-related employment". This includes the following industries;

```{r ag-related industry list}
agrelated.ind.list <- read_csv("Data/Earnings/Master-ag-related-ind-wages-pr.csv") %>%
  distinct(naicstitle) %>%
  rename(`Ag-related NAICS categories` = 1)

kable(agrelated.ind.list, format = "html", align = "c") %>%
    kable_styling(bootstrap_options = "striped") %>%
    column_spec(1, width = "10em")

```

<br>

# Analysis

The purpose of this analysis was to determine where the share of the workforce in each ag-related industry was located, what their wage per employment was, and which industries were driving these wages. This is done through the following sections.

**Wages per employment of ag-related industries combined**: This section combines all of the wages of all the ag-related industries together and divides by the total employment for these industries to provide a wage per employment value. We will then assess which regions have the highest values.

**Ag-related industry with the highest wage per employment and number of employed**: This section provides the wage per employment and the number of employed for each of the 9 ag-related industries across the entire state. We will then assess which industries seem to drive employment and have the highest wages.

**Concentration of ag-related industries**": This section provides the wage per employment for the ag-related industries in each region and also the share of the employment that each region employs. We will then assess any differences in the share of employment as well as the difference in wages of those industries.

<br>

## Wages per employment of ag-related industries combined {.tabset}

It's clear that the highest wages per employment are located in our most metropolitan counties in the state. In fact, the highest wages per employment are located in Hennepin and Ramsey Counties. Wages per employment in these counties and other entirely urban counties across the state are between 30% and 50% higher than the rest of the state.

In Southwest Minnesota, the highest wages are clustered in and around Blue Earth County and Lyons County. In the Northwest, the wages are typically a bit higher than Southwest. Perhaps this is due to the different crops being marketed and sold?

<br>

```{r prep ag wages per emp combined, include=FALSE, cache=TRUE}

ag.wages.emp.comb.ruca <- ag.wages.emp.ruca %>%
  group_by(periodyear, Dem_Desc) %>%
  summarise(ag.wages = sum(wages[naicstitle != "Total, All Industries"]),
            total.wages = sum(wages[naicstitle == "Total, All Industries"]),
            ag.emp = sum(emp[naicstitle != "Total, All Industries"]),
            total.emp = sum(emp[naicstitle == "Total, All Industries"])) %>%
  ungroup() %>%
  mutate(ag.wages.per.emp = ag.wages / ag.emp,
         total.wages.per.emp = total.wages / total.emp)

ag.wages.emp.comb.pr <- ag.wages.emp.pr %>%
  group_by(periodyear, planning.region) %>%
  summarise(ag.wages = sum(wages[naicstitle != "Total, All Industries"]),
            total.wages = sum(wages[naicstitle == "Total, All Industries"]),
            ag.emp = sum(emp[naicstitle != "Total, All Industries"]),
            total.emp = sum(emp[naicstitle == "Total, All Industries"])) %>%
  ungroup() %>%
  mutate(ag.wages.per.emp = ag.wages / ag.emp,
         total.wages.per.emp = total.wages / total.emp)

ag.wages.emp.comb.edr <- ag.wages.emp.edr %>%
  group_by(periodyear, edr, planning.region) %>%
  summarise(ag.wages = sum(wages[naicstitle != "Total, All Industries"]),
            total.wages = sum(wages[naicstitle == "Total, All Industries"]),
            ag.emp = sum(emp[naicstitle != "Total, All Industries"]),
            total.emp = sum(emp[naicstitle == "Total, All Industries"])) %>%
  ungroup() %>%
  mutate(ag.wages.per.emp = ag.wages / ag.emp,
         total.wages.per.emp = total.wages / total.emp)

ag.wages.emp.comb.county <- ag.wages.emp.county %>%
  group_by(periodyear, areaname, countyfp) %>%
  summarise(ag.wages = sum(wages[naicstitle != "Total, All Industries"]),
            total.wages = sum(wages[naicstitle == "Total, All Industries"]),
            ag.emp = sum(emp[naicstitle != "Total, All Industries"]),
            total.emp = sum(emp[naicstitle == "Total, All Industries"])) %>%
  ungroup() %>%
  mutate(ag.wages.per.emp = ag.wages / ag.emp,
         total.wages.per.emp = total.wages / total.emp,
         ag.wages.per.emp = ifelse(areaname == "Mahnomen", NA, ag.wages.per.emp)) %>%
  left_join(mn_counties[,c(4,7)], by = c("countyfp" = "FIPS_CODE"))

```


### RUCA

It's pretty clear to see that the highest paying jobs in ag-related industries are located in our entirely urban counties. By 2018, wages in ag-related industries were $66,862 per employment. This was 32% to 45% more than the other county groups.

<br>

```{r chart ag wages per employment combined ruca}

ag.wages.emp.comb.ruca.plot <- ggplot(ag.wages.emp.comb.ruca, aes(periodyear, ag.wages.per.emp, color = Dem_Desc)) +
  geom_smooth(se = FALSE, size = 2) +
  geom_point_interactive(size = 3, aes(data_id = ag.wages.per.emp, tooltip = paste(Dem_Desc, "\nYear: ", periodyear, "\nTotal wages in ag-related industries: ", dollar(ag.wages), "\nTotal employment in ag-related industries: ", comma(ag.emp), "\nAg wages per employment: ", dollar(ag.wages.per.emp), sep = ""))) +
  labs(x="", y = "", color="", title = "Ag wages per employment")+
  scale_y_continuous(labels=scales::dollar)+
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_manual(values= color.ruca,
                     guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom")

girafe(ggobj = ag.wages.emp.comb.ruca.plot, width_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```

### Planning Region

When broken down by planning regions, the seven county metro has the highest paying jobs. The $71,868 earned there is about 38% to 79% higher. Interestingly, the Southwest region has the 2nd highest wages per employment even though it has a large number of entirely rural counties. What this indicates is that the more urban counties in the region are earning significantly more wages than the rural counties, which drives up the wages in that region. So the disparities are still between urban-ness and rural-ness.

<br>

```{r chart ag wages per employment combined pr}

ag.wages.emp.comb.pr.plot <- ggplot(ag.wages.emp.comb.pr, aes(periodyear, ag.wages.per.emp, color = planning.region)) +
  geom_smooth(se = FALSE, size = 2) +
  geom_point_interactive(size = 3, aes(data_id = ag.wages.per.emp, tooltip = paste(planning.region, "\nYear: ", periodyear, "\nTotal wages in ag-related industries: ", dollar(ag.wages), "\nTotal employment in ag-related industries: ", comma(ag.emp), "\nAg wages per employment: ", dollar(ag.wages.per.emp), sep = ""))) +
  labs(x="", y = "", color="", title = "Ag wages per employment")+
  scale_y_continuous(labels=scales::dollar)+
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_manual(values= color.pr,
                     guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom")

girafe(ggobj = ag.wages.emp.comb.pr.plot, width_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```


### EDR

The breakdown by EDR shows that EDRs in Southwest typically have higher wages per employment when there are larger metros within the edr. For example, EDR 9 - South Central has Blue Earth County, which is an entirely urban county. This EDR also has the highest wages per employment in that region.

A similar situation is in the Central region where EDR 6E and 7W have the highest wages. These contain counties such as Stearns and Kandiyohi which are considered more urban than other counties in the planning region.

The Northwest is a bit different due to the varying ecosystem in that planning region. There, we would expect EDR 2 - Headwaters to be lower since that region doesn't typically have agriculture as a primary activity nearby.

<br>

```{r chart ag wages per employment combined edr, fig.height=9}

ag.wages.emp.comb.edr.plot <- ggplot(ag.wages.emp.comb.edr, aes(periodyear, ag.wages.per.emp, color = edr)) +
  facet_wrap(~planning.region, ncol = 2) +
  geom_smooth(se = FALSE, size = 2) +
  geom_point_interactive(size = 3, aes(data_id = ag.wages.per.emp, tooltip = paste(edr, "\nYear: ", periodyear, "\nTotal wages in ag-related industries: ", dollar(ag.wages), "\nTotal employment in ag-related industries: ", comma(ag.emp), "\nAg wages per employment: ", dollar(ag.wages.per.emp), sep = ""))) +
  labs(x="", y = "", color="", title = "Ag wages per employment")+
  scale_y_continuous(labels=scales::dollar)+
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_manual(values= color.edr,
                     guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 10))

girafe(ggobj = ag.wages.emp.comb.edr.plot, height_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```

### 2018 county map

This map highlights the concentration of the highest paying jobs are located in the seven-county metro, Blue Earth County and Martin County. 

In the Southwest region, the highest wages per employment are clustered in and around Lyons County, including Yellow Medicine, Redwood Falls, and Lac qui Parle counties.
 
Up in the Northwest region, the highest wages per employment are more spread out. It goes from Stevens and Grant Counties and then directly along the western border, skipping over Norman, and seeing high wages North of there.

<br>

```{r map 2018 ag wages per employment}
ag.wages.emp.comb.county.plot <- ggplot(filter(ag.wages.emp.comb.county, periodyear == 2018)) +
  geom_sf_interactive(aes(fill = ag.wages.per.emp, tooltip = paste(areaname, "\nYear: ", periodyear, "\nTotal wages in ag-related industries: ", dollar(ag.wages), "\nTotal employment in ag-related industries: ", comma(ag.emp), "\nWages per employment: ", dollar(ag.wages.per.emp), sep = ""), data_id = areaname, geometry = geometry)) +
  scale_fill_gradient(low = "white",
                      high = "#076324")+
  theme_sf+
  labs(title="2018 ag wages per employment") +
  theme(text = element_text(size = 10))

girafe(ggobj = ag.wages.emp.comb.county.plot, width_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))


```

<br>

## Ag-related industry with highest wage per employment and number of employed

The two merchant wholesaler industries have the highest wages per employment by far compared to the other industries while also employing a modest number. Food manufacturing is by the highest number of employed, but is also one of the lowest paying per employment.

One important aspect to keep in mind that some of these industries likely employ only part-time of the year. For example, support activities for crop production likely only employ during planting and harvesting times of the year. Thus, their wages earned are going to be significantly lower.

<br>

```{r prep each industry wage per employment statewide, include=FALSE, cache=TRUE}

ag.wages.per.emp.ind.mn <- read_csv("Data/Employment/qcew-ag-related-industries-mn.csv") %>%
  select(1,2,5,28,30) %>%
  rename(emp = empYear,
         wages = totwageYear) %>%
  group_by(periodyear, naicstitle) %>%
  summarise(wages = sum(wages, na.rm = TRUE),
            emp = sum(emp, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(wages.per.emp = wages / emp)
```

```{r chart each industry wages per employment statewide}

ag.wages.per.emp.ind.mn.plot <- ggplot(filter(ag.wages.per.emp.ind.mn, periodyear == 2018), aes(emp, wages.per.emp, color = naicstitle)) +
  geom_label_repel(size = 2, aes(x = emp, y = wages.per.emp, label = naicstitle), show.legend = FALSE) +
  geom_point_interactive(size = 2, aes(data_id = wages.per.emp, tooltip = paste("NAICS: ", naicstitle, "\nNumber employed: ", comma(emp), "\nWages per employed: ", dollar(wages.per.emp), sep = "")), show.legend = FALSE) +
  labs(x="", y = "", color="", title = "Wages per employment and number employed in ag-related\nindustries")+
  scale_y_continuous(labels=scales::dollar)+
  scale_x_continuous(labels=scales::comma) +
  theme_line+
  scale_color_discrete(guide = guide_legend(ncol = 2)) +
  theme(legend.position = "bottom")

girafe(ggobj = ag.wages.per.emp.ind.mn.plot, width_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))


```

<br>

## Concentration of ag-related industries {.tabset}

Due to the shear size of the workforce in our most urban areas, they also have the highest concentration of workforce for each of the ag-related industries. However, breaking it down by planning region shows that it isn’t necessarily the seven-county metro that has a higher percentage of the workforce across all these industries, but rather, there are metropolitan counties in the other regions that also share a considerable amount of the workforce. 

Not surprisingly, the Southwest, Northwest and Central regions of the state typically have the highest shares of the workforce in each industry. However, the seven county metro has the highest share of the workforce in farm raw product material merchant wholesalers and the farm supplies merchant wholesalers by a significant margin. In addition, the wages per employment are significantly higher within this industry in the seven-county metro compared to the other regions. In fact, it's about double.

The EDR-level shows that the share of each industry's workforce within each planning region is shared relatively equally across all the EDRs.

<br>

```{r prep concentration of ag-related industries, include=FALSE, cache=TRUE}

ag.ind.conc.ruca <- ag.wages.emp.ruca %>%
  group_by(periodyear, naicstitle) %>%
  mutate(pct.ind.emp = emp / sum(emp)) %>%
  ungroup()

ag.ind.conc.pr <- ag.wages.emp.pr %>%
  group_by(periodyear, naicstitle) %>%
  mutate(pct.ind.emp = emp / sum(emp)) %>%
  ungroup()

ag.ind.conc.edr <- ag.wages.emp.edr %>%
  group_by(periodyear, naicstitle) %>%
  mutate(pct.ind.emp = emp / sum(emp)) %>%
  ungroup() %>%
  mutate(label = str_sub(edr, 1,6),
         label = str_trim(label, side = "both"))

```

### RUCA

Due to the size of the shear workforce located in our entirely urban counties, they have a majority of the workforce in each of these industries except for animal production and support activities for crop production. 

For the most part, wages per employment are similar across all of the regions except for ag implement manufacturing, farm product raw material merchant wholesalers, and farm supplies merchant wholesalers where the values are significantly higher than other county group types. 

One other interesting element is that our entirely rural county group has a very low percentage of the workforce in each of these industries except for support activities for crop production where it has 40% of the total workforce in that industry.

<br>

```{r chart ag ind concentration, fig.height=8}
ag.ind.conc.ruca.plot <- ggplot(filter(ag.ind.conc.ruca, periodyear == 2018, naicstitle != "Total, All Industries"), aes(pct.ind.emp, wages.emp, color = Dem_Desc)) +
  facet_wrap(~naicstitle, ncol = 2) +
  geom_smooth(se = FALSE, size = 2) +
  geom_point_interactive(size = 3, aes(data_id = pct.ind.emp, tooltip = paste(Dem_Desc, "\nNAICS: ", naicstitle, "\nEmployment in industry: ", comma(emp), "\nTotal wages earned in industry: ", dollar(wages), "\nWages per employment: ", dollar(wages.emp), "\nPercent of industry employment located in this region: ", percent(pct.ind.emp), sep = "")), show.legend = FALSE) +
  geom_text_repel(size = 3, aes(label = Dem_Desc), show.legend = FALSE) +
  labs(x="", y = "", color="", title = "Percent of industry employment located in each region by\nwage per employment earned")+
  scale_y_continuous(labels=scales::dollar)+
  scale_x_continuous(labels=scales::percent) +
  theme_line+
  scale_color_manual(values= color.ruca,
                     guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 12))

girafe(ggobj = ag.ind.conc.ruca.plot, height_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```

### Planning Region

Breaking it down by planning region shows that it isn't, necessarily, the seven-county metro that has a higher percentage of the workforce across all these industries, but rather, there are metropolitan counties in the other regions that also share a considerable amount of the workforce. There are many industries where the seven county metro, which is comprised of entirely urban counties, does not have the highest share of employment.

Here's a breakdown of each industry and location of the most significant share of the workforce.

Ag implement manufacturing: A majority of the workforce is in the Central region with 54%. Northwest also has a significant amount with 29%. The wage per employment is significantly higher in the seven-county metro.

Animal production: A significant portion of the workforce is located in Southwest with 35%, followed by Central with 25%. Wages per employment are even across all regions.

Crop production: Interestingly, the seven-county metro has the highest share of this industry with 33% followed closely by Northwest with 30%. Wages per employment are even across all regions.

Farm product raw material merchant wholesalers: The Seven-county metro has the highest share by far with 44% of the workforce. The next closest is Southwest with 26% and Northwest with 17%. Huge disparities in wages per employment in this industry with the seven county metro having the highest value of $154,727. The next closest is in Northwest which is 50% of that value - $62,424.

Farm supplies merchant wholesalers: Again, the seven-county metro has the highest share with 30% of the workforce. Southwest isn't too far behind with 26%. There are disparities in wages per employment here as well but not as much. The seven county metro has $118,694 per employment while the next closest is Southwest with $72,473 per employment.

Food manufacturing: Except for the Northeast, that share of food manufacturing employment is spread pretty evenly across all the regions with shares ranging from 13% in Northwest to 23% in the Southwest. Wages per employment are even across all regions.

Fertilizer and chemical manufacturing: The seven county metro leads the way with 61% of the employment in this industry followed by Southwest with 39%. This is a relatively small piece of the ag-related employment puzzle. Wages per employment are about $20,000 per employment higher in the seven-county metro than in Southwest.

Support activities for animal production: The seven county metro, Southwest, and Central all have a similar share of the workforce with about 25% each. Wages are relatively similar.

Support activities for crop production: Northwest has the highest share of the workforce with 39% followed by Southwest with 26%. Wages per employment are relatively even across all the regions.

<br>

```{r chart ag ind concentration pr, fig.height=8}
ag.ind.conc.pr.plot <- ggplot(filter(ag.ind.conc.pr, periodyear == 2018, naicstitle != "Total, All Industries"), aes(pct.ind.emp, wages.emp, color = planning.region)) +
  facet_wrap(~naicstitle, ncol = 2) +
  geom_smooth(se = FALSE, size = 2) +
  geom_point_interactive(size = 3, aes(data_id = pct.ind.emp, tooltip = paste(planning.region, "\nNAICS: ", naicstitle, "\nEmployment in industry: ", comma(emp), "\nTotal wages earned in industry: ", dollar(wages), "\nWages per employment: ", dollar(wages.emp), "\nPercent of industry employment located in this region: ", percent(pct.ind.emp), sep = "")), show.legend = FALSE) +
  geom_text_repel(size = 3, aes(label = planning.region), show.legend = FALSE) +
  labs(x="", y = "", color="", title = "Percent of industry employment located in each region by\nwage per employment earned")+
  scale_y_continuous(labels=scales::dollar)+
  scale_x_continuous(labels=scales::percent) +
  theme_line+
  scale_color_manual(values= color.pr,
                     guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 12))

girafe(ggobj = ag.ind.conc.pr.plot, height_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```

### EDR

This level provides a bit more detail to help us identify which EDRs have some of the concentration of industries in each planning region.

For example, we can see that animal production employment is concentrated in EDR 9, which explains why Southwest had such a high concentration in the planning region breakdown.

Interestingly, we can see that when broken up by EDRs, the employment in the two merchant wholesaler industries is really spread evenly throughout the Northwest and Southwest regions. Of course, the seven county metro continues to the highest in each of those industries.

For the most part, employment in the industries is spread out pretty evenly across the EDRs in each planning region, which is why they are clustered so closely together.

<br>

```{r chart ag ind concentration edr, fig.height=8}
ag.ind.conc.edr.plot <- ggplot(filter(ag.ind.conc.edr, periodyear == 2018, naicstitle != "Total, All Industries"), aes(pct.ind.emp, wages.emp, color = label), show.legend = FALSE) +
  facet_wrap(~naicstitle, ncol = 2) +
  geom_smooth(se = FALSE, size = 2) +
  geom_point_interactive(size = 3, aes(data_id = pct.ind.emp, tooltip = paste(edr, "\n Planning region: ", planning.region, "\nNAICS: ", naicstitle, "\nEmployment in industry: ", comma(emp), "\nTotal wages earned in industry: ", dollar(wages), "\nWages per employment: ", dollar(wages.emp), "\nPercent of industry employment located in this region: ", percent(pct.ind.emp), sep = "")), show.legend = FALSE) +
  geom_text_repel(size = 3, aes(label = label), show.legend = FALSE) +
  labs(x="", y = "", color="", title = "Percent of industry employment located in each region by wage per employment earned")+
  scale_y_continuous(labels=scales::dollar)+
  scale_x_continuous(labels=scales::percent) +
  theme_line+
  scale_color_manual(values= color.label.edr,
                     guide = guide_legend(ncol = 3)) +
  theme(legend.position = "none",
        text = element_text(size = 12))

girafe(ggobj = ag.ind.conc.edr.plot, height_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))
```

<br>

```{r prep ag wages and employment by industry, include=FALSE, cache=TRUE}

ag.wages.emp.pct.ruca <- ag.wages.emp.ruca %>%
  group_by(periodyear, Dem_Desc) %>%
  mutate(pct.emp = emp / sum(emp[naicstitle != "Total, All Industries"])) %>%
  ungroup()

ag.wages.emp.pct.pr <- ag.wages.emp.pr %>%
  group_by(periodyear, planning.region) %>%
  mutate(pct.emp = emp / sum(emp[naicstitle != "Total, All Industries"])) %>%
  ungroup()

ag.wages.emp.pct.edr <- ag.wages.emp.edr %>%
  group_by(periodyear, edr, planning.region) %>%
  mutate(pct.emp = emp / sum(emp[naicstitle != "Total, All Industries"])) %>%
  ungroup()

ag.wages.emp.pct.county <- ag.wages.emp.county %>%
  group_by(periodyear, areaname, countyfp) %>%
  mutate(pct.emp = emp / sum(emp[naicstitle != "Total, All Industries"])) %>%
  ungroup() %>%
  left_join(mn_counties[,c(4,7)], by = c("countyfp" = "FIPS_CODE"))


```


## County map of most prominent ag-related industry - 2018

The maps below provide the highest and 2nd highest ag-related industry in each county.

Food production is obviously a major employment industry in agriculture. It's in the top 2 highest employment industry for a large majority of counties across the state and makes up 50% or more of ag-related employment across nearly all of these counties.

The two merchant wholesaler industries pop up in Hennepin and Ramsey County as well as in a large contingent of counties in Southwest Minnesota. It's also prominent in a few counties in NW Minnesota. 

<br>

```{r prep industry employment rank, include=FALSE, cache=TRUE}

ag.ind.rank.county <- ag.wages.emp.pct.county %>%
  filter(periodyear == 2018) %>%
  group_by(areaname, countyfp) %>%
  arrange(desc(pct.emp)) %>%
  mutate(rank = seq(n())) %>%
  ungroup() %>%
  select(2,3,6,5,11,12,13,15) %>%
  complete(rank, countyfp) %>%
  group_by(countyfp) %>%
  fill(areaname) %>%
  ungroup() %>%
  left_join(mn_counties[,c(4,7)], by = c("countyfp" = "FIPS_CODE")) %>%
  mutate(naicstitle = as.factor(naicstitle),
         rank = rank - 1)

```

```{r maps industry employment rank, fig.height=7}

ag.ind.rank.county.1.plot <- ggplot(filter(ag.ind.rank.county, rank == 1)) +
  geom_sf_interactive(aes(fill = naicstitle, tooltip = paste(areaname, "\nNAICS: ", naicstitle, "\nRank: ", rank, "\nTotal wages earned: ", dollar(wages), "\nEmployment: ", comma(emp), "\nPercent of ag-related employment: ", percent(pct.emp), "\nWages per employment: ", dollar(wages.emp), sep = ""), data_id = areaname, geometry = geometry)) +
  scale_fill_manual(values = c("Crop Production" = "#a6cee3", "Animal Production and Aquaculture" = "#1f78b4", "Support Activities for Crop Production" = "#b2df8a", "Support Activities for Animal Production" = "#33a02c", "Food Manufacturing" = "#fb9a99", "Agricultural Implement Manufacturing" = "#e31a1c", "Farm Product Raw Material Merchant Wholesalers" = "#fdbf6f", "Farm Supplies Merchant Wholesalers" = "#ff7f00", "Pesticide, Fertilizer, and Other Agricultural Chemical Manufacturing" = "#cab2d6", "Total, All Industries" = "black", "NA" = "white"),
                    guide = guide_legend(ncol = 2),
                    drop = FALSE,
                    breaks = c("Crop Production", "Animal Production and Aquaculture", "Support Activities for Crop Production", "Support Activities for Animal Production", "Food Manufacturing", "Agricultural Implement Manufacturing", "Farm Product Raw Material Merchant Wholesalers", "Farm Supplies Merchant Wholesalers", "Pesticide, Fertilizer, and Other Agricultural Chemical Manufacturing"))+
  theme_sf+
  labs(title="Ag-industry with 2nd highest employment") +
  theme(text = element_text(size = 10),
        legend.position = "bottom")

ag.ind.rank.county.2.plot <- ggplot(filter(ag.ind.rank.county, rank == 2)) +
  geom_sf_interactive(aes(fill = naicstitle, tooltip = paste(areaname, "\nNAICS: ", naicstitle, "\nRank: ", rank, "\nTotal wages earned: ", dollar(wages), "\nEmployment: ", comma(emp), "\nPercent of ag-related employment: ", percent(pct.emp), "\nWages per employment: ", dollar(wages.emp), sep = ""), data_id = areaname, geometry = geometry)) +
  scale_fill_manual(values = color.ag.ind,
                    guide = guide_legend(ncol = 2))+
  theme_sf+
  labs(title="Ag-industry with highest employment") +
  theme(text = element_text(size = 10),
        legend.position = "bottom")

ag.ind.rank.county.1.plot.legend <- get_legend(ag.ind.rank.county.1.plot + theme(legend.box.margin = margin(0,0,0,0)))

ag.ind.rank.county.1.plot.legend.bottom <- plot_grid(NULL, ag.ind.rank.county.1.plot.legend, NULL, ncol = 3, rel_widths = c(0,.5,3))


girafe(ggobj = plot_grid(ag.ind.rank.county.1.plot + theme(legend.position = "none"), ag.ind.rank.county.2.plot + theme(legend.position = "none"), NULL, ag.ind.rank.county.1.plot.legend.bottom, ncol = 2, rel_heights = c(1,.4)), width_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```


