---
title: "Land values vs. net income"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
runtime: shiny
resource_files:
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.cpg
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.dbf
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.prj
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbn
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbx
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shp.xml
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
```

```{r loading jon docs and shapefiles}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", size = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.title = element_blank(),
        text = element_text(size = 15),
        plot.caption = element_text(hjust = 0.5, face = "italic"),
        legend.text = element_text(margin = margin(l = 2, r = 5)))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", size = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        text = element_text(size = 15),
        plot.caption = element_text(hjust = 0.5, face = "italic"),
        legend.text = element_text(margin = margin(l = 2, r = 5)))

theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        plot.title = element_text(hjust = 0.5),
        legend.title = element_blank(),
        text = element_text(size = 15),
        plot.caption = element_text(hjust = 0.5, face = "italic"),
        legend.text = element_text(margin = margin(l = 2, r = 5)))

counties.regions <- read_csv("Data/Join docs/county_regions.csv") %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         edr = str_replace(edr, "  ", " "))

color.ruca <- c("Entirely rural" = "#5CA81F", "Town/rural mix" = "#C7EF99", "Urban/town/rural mix" = "#d8b365", "Entirely urban" = "#a6611a", "Minnesota" = "black")

color.pr <- c("Northwest" = "#810f7c","Northeast" = "#fe9929", "Central" = "#076324", "Seven County Mpls-St Paul" = "#d8b365", "Southwest" = "#1f78b4", "Southeast" = "#d7301f", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.counties <- scale_color_brewer(palette = "Dark2",
                       guide = guide_legend(ncol = 3))

mn_counties <- st_read("Data/Shapefiles/county shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE)
```

```{r master land value and net income objects, include=FALSE, cache=TRUE}
master.netinc.county <- read_csv("Data/Income and expenses/Master-farm-income-expenses.csv") %>%
  mutate(Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = str_replace(edr, "  ", " "),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota")) %>%
  left_join(counties.regions[,c(1,2)], by = "countyfp") %>%
  filter(Name != "Minnesota")

master.lv.county <- read_csv("Data/Land values/Master-ag-land-values.csv") %>%
  mutate(Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = str_replace(edr, "  ", " "),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota")) %>%
  rename(year = Year) %>%
  filter(County != "Minnesota") %>%
  rename(Name = County)

```

<br>

# Summary

**Definition of net income**: Net income is calculated by adding the cash receipts from marketings to other income and subtracting the production expenses.

**Definition of cash receipts from marketings**: consists of the gross revenue received by farmers from the sale of crops, livestock, and livestock products and of the value of defaulted loans made by Commodity Credit Corporation (CCC) and secured by crops.

**Definition of other income**: comprised of Government Payments and Imputed and miscellaneous income received.

**Definition of production expenses**: Farm production expenses consist of purchases of feed, livestock and poultry, seed, fertilizer, agricultural chemicals and lime, and petroleum products; labor expenses; machinery rental and custom work; animal health costs; and all other expenses including depreciation. BEA adjusts the USDA state estimates of production expenses to account for methodological differences in the treatment of depreciation and to conform to BEA definitions and classifications.

**Definition of class 2a agricultural land**: The estimated land values are taking from the University of Minnesota - Land Economics program. The estimates used are based estimates provided by county assessors and is for Class 2a agricultural land. This is all land that can be cultivated for row crops or grains.

<br>

# Analysis

[Analysis description here]

<br>

```{r prep net income and land values index, include=FALSE, cache=TRUE}
net.income.acre.ruca <- master.netinc.county %>%
  group_by(Dem_Desc, year) %>%
  summarise(net.income = sum(net.income, na.rm = TRUE),
            total.acres = sum(Total.acres, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(net.income.acre = net.income / total.acres) %>%
  group_by(Dem_Desc) %>%
  mutate(net.income.acre.index = net.income.acre / net.income.acre[year == 2001]) %>%
  ungroup()

land.value.acre.ruca <- master.lv.county %>%
  group_by(Dem_Desc, year) %>%
  summarise(total.acres = sum(Total.acres, na.rm = TRUE),
            total.value = sum(Total.estimated.value, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(land.value.acre = total.value / total.acres) %>%
  group_by(Dem_Desc) %>%
  mutate(land.value.acre.index = land.value.acre / land.value.acre[year == 2001]) %>%
  ungroup()

net.income.acre.pr <- master.netinc.county %>%
  group_by(planning.region, year) %>%
  summarise(net.income = sum(net.income, na.rm = TRUE),
            total.acres = sum(Total.acres, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(net.income.acre = net.income / total.acres) %>%
  group_by(planning.region) %>%
  mutate(net.income.acre.index = net.income.acre / net.income.acre[year == 2001]) %>%
  ungroup()

land.value.acre.pr <- master.lv.county %>%
  group_by(planning.region, year) %>%
  summarise(total.acres = sum(Total.acres, na.rm = TRUE),
            total.value = sum(Total.estimated.value, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(land.value.acre = total.value / total.acres) %>%
  group_by(planning.region) %>%
  mutate(land.value.acre.index = land.value.acre / land.value.acre[year == 2001]) %>%
  ungroup()

net.income.acre.edr <- master.netinc.county %>%
  group_by(edr, planning.region, year) %>%
  summarise(net.income = sum(net.income, na.rm = TRUE),
            total.acres = sum(Total.acres, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(net.income.acre = net.income / total.acres) %>%
  group_by(edr, planning.region) %>%
  mutate(net.income.acre.index = net.income.acre / net.income.acre[year == 2001]) %>%
  ungroup()

land.value.acre.edr <- master.lv.county %>%
  group_by(edr, planning.region, year) %>%
  summarise(total.acres = sum(Total.acres, na.rm = TRUE),
            total.value = sum(Total.estimated.value, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(land.value.acre = total.value / total.acres) %>%
  group_by(planning.region) %>%
  mutate(land.value.acre.index = land.value.acre / land.value.acre[year == 2001]) %>%
  ungroup()

```

## Net income from farming per acre vs. land value per acre {.tabset}

When comparing the net income from farming and the land values, the most striking trend is that land values significantly increased along with net income during the late 2000’s and early 2010’s. However, now that net income is significantly lower, below breaking even in some places, land values have barely decreased. Interestingly, land values in our more metro areas of the state have stayed significantly higher than traditional agricultural regions due to residential and commercial development pressures.

The highest values, outside of the seven county metro, are found in the Southwest and Southeast parts of the state. Surprisingly, land values in Northwest Minnesota have remained significantly lower than other parts of the state.

<br>

### RUCA

The trend of net incomes for farms across the county groups is not surprising. It's pretty easy to see where the crop commodity boom began and ended. Every region across the state saw a spike in farm land values during this time frame. Land values are beginning to drop slightly across the state but nowhere near the rate of decline experienced in net incomes. In addition, the Entirely urban counties aren't really seeing much of a decline, likely due to those land values being driven by commercial and residential development pressure.

<br>

```{r chart net income and land value index ruca, fig.width=8}

net.income.acre.ruca.plot <- ggplot(net.income.acre.ruca, aes(year, net.income.acre, color = Dem_Desc)) +
  geom_smooth(se = FALSE, size = 1) +
  geom_hline(yintercept = 0, color = "black") +
  geom_point_interactive(size = 2, aes(data_id = net.income.acre.index, tooltip = paste(Dem_Desc, "\nYear: ", year, "\nTotal farm net income: ", dollar(net.income), "\nTotal acres: ", comma(total.acres), "\nNet income per acre: ", dollar(net.income.acre), "\nNet income per acre indexed to 2001 levels: ", percent(net.income.acre.index), sep = ""))) +
  labs(x="", y = "", color="", title = "Net income per farm acre")+
  scale_y_continuous(labels=scales::dollar)+
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_manual(values = color.ruca,
                     guide = guide_legend(ncol = 4)) +
  theme(legend.position = "bottom",
        text = element_text(size = 10))

land.value.acre.ruca.plot <- ggplot(land.value.acre.ruca, aes(year, land.value.acre, color = Dem_Desc)) +
  geom_smooth(se = FALSE, size = 1) +
  geom_hline(yintercept = 0, color = "black") +
  geom_point_interactive(size = 2, aes(data_id = land.value.acre, tooltip = paste(Dem_Desc, "\nYear: ", year, "\nTotal acres: ", comma(total.acres), "\nTotal estimated land value: ", dollar(total.value),  "\nLand value per acre: ", dollar(land.value.acre), "\nLand value per acre indexed to 2001 levels: ", percent(land.value.acre.index), sep = ""))) +
  labs(x="", y = "", color="", title = "Land value per farm acre")+
  scale_y_continuous(labels=scales::dollar)+
  coord_cartesian(ylim = seq(-250,7000, 200)) +
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_manual(values = color.ruca,
                     guide = guide_legend(ncol = 4)) +
  theme(legend.position = "bottom",
        text = element_text(size = 10))

net.income.acre.ruca.plot.legend <- get_legend(net.income.acre.ruca.plot + theme(legend.box.margin = margin(0,0,0,0)))

net.income.acre.ruca.plot.legend.bottom <- plot_grid(NULL, net.income.acre.ruca.plot.legend, NULL, ncol = 3, rel_widths = c(0,.5,3))

girafe(ggobj = plot_grid(net.income.acre.ruca.plot + theme(legend.position = "none"), land.value.acre.ruca.plot + theme(legend.position = "none"), NULL, net.income.acre.ruca.plot.legend.bottom, ncol = 2, rel_heights = c(1, .1)), width_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```

### Planning Region

Regions where land values seem to be reacting the most to farm net incomes are the Southwest and Southeast. Interestingly, land in the Northwest isn't as high as expected. But maybe we will see more nuance there when broken down by EDR.

In the Southwest region, land values peaked in 2014 at $7,670 per acre and have dropped over $1,000 to $6,365. 

The Southeast region had land values peak in 2014 as well at $6,642 and have dropped over $500 per acre to $6,091.

<br>

```{r chart net income and land value index pr, fig.width=8}

net.income.acre.pr.plot <- ggplot(net.income.acre.pr, aes(year, net.income.acre, color = planning.region)) +
  geom_smooth(se = FALSE, size = 1) +
  geom_hline(yintercept = 0, color = "black") +
  geom_point_interactive(size = 2, aes(data_id = net.income.acre.index, tooltip = paste(planning.region, "\nYear: ", year, "\nTotal farm net income: ", dollar(net.income), "\nTotal acres: ", comma(total.acres), "\nNet income per acre: ", dollar(net.income.acre), "\nNet income per acre indexed to 2001 levels: ", percent(net.income.acre.index), sep = ""))) +
  labs(x="", y = "", color="", title = "Net income per farm acre")+
  scale_y_continuous(labels=scales::dollar)+
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_manual(values = color.pr,
                     guide = guide_legend(ncol = 4)) +
  theme(legend.position = "bottom",
        text = element_text(size = 10))

land.value.acre.pr.plot <- ggplot(land.value.acre.pr, aes(year, land.value.acre, color = planning.region)) +
  geom_smooth(se = FALSE, size = 1) +
  geom_hline(yintercept = 0, color = "black") +
  geom_point_interactive(size = 2, aes(data_id = land.value.acre, tooltip = paste(planning.region, "\nYear: ", year, "\nTotal acres: ", comma(total.acres), "\nTotal estimated land value: ", dollar(total.value),  "\nLand value per acre: ", dollar(land.value.acre), "\nLand value per acre indexed to 2001 levels: ", percent(land.value.acre.index), sep = ""))) +
  labs(x="", y = "", color="", title = "Land value per farm acre")+
  scale_y_continuous(labels=scales::dollar)+
  coord_cartesian(ylim = seq(-3250,13000, 2500)) +
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_manual(values = color.pr,
                     guide = guide_legend(ncol = 4)) +
  theme(legend.position = "bottom",
        text = element_text(size = 10))

net.income.acre.pr.plot.legend <- get_legend(net.income.acre.pr.plot + theme(legend.box.margin = margin(0,0,0,0)))

net.income.acre.pr.plot.legend.bottom <- plot_grid(NULL, net.income.acre.pr.plot.legend, NULL, ncol = 3, rel_widths = c(0,.5,3))

girafe(ggobj = plot_grid(net.income.acre.pr.plot + theme(legend.position = "none"), land.value.acre.pr.plot + theme(legend.position = "none"), NULL, net.income.acre.pr.plot.legend.bottom, ncol = 2, rel_heights = c(1, .1)), width_svg = 9) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```

### EDR

In the Northwest region, the trend line for land values in EDR 4 - West Central indicates some decline over the last few years and are likely being impacted by falling net incomes.

In Central MN, land values in EDR 6E - Southwest Central have declined over the last few years indicating impacts from falling net farm incomes.

All the EDRs in Southwest MN show declines in land values over the last few years. This is also where the highest land values are found among agriculture regions.



<br>

```{r chart net income and land value index edr, fig.height=10}

net.income.acre.edr.plot <- ggplot(net.income.acre.edr, aes(year, net.income.acre, color = edr)) +
  geom_smooth(se = FALSE, size = 1) +
  facet_wrap(~planning.region, ncol = 1) +
  geom_hline(yintercept = 0, color = "black") +
  geom_point_interactive(size = 2, aes(data_id = net.income.acre.index, tooltip = paste(edr, "\nYear: ", year, "\nTotal farm net income: ", dollar(net.income), "\nTotal acres: ", comma(total.acres), "\nNet income per acre: ", dollar(net.income.acre), "\nNet income per acre indexed to 2001 levels: ", percent(net.income.acre.index), sep = ""))) +
  labs(x="", y = "", color="", title = "Net income per farm acre")+
  scale_y_continuous(labels=scales::dollar)+
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_manual(values = color.edr,
                     guide = guide_legend(ncol = 4)) +
  theme(legend.position = "bottom",
        text = element_text(size = 10),
        legend.text = element_text(size = 6),
        axis.text.x = element_text(size = 6))

land.value.acre.edr.plot <- ggplot(land.value.acre.edr, aes(year, land.value.acre, color = edr)) +
  geom_smooth(se = FALSE, size = 1) +
  facet_wrap(~planning.region, ncol = 1) +
  geom_hline(yintercept = 0, color = "black") +
  geom_point_interactive(size = 2, aes(data_id = land.value.acre, tooltip = paste(edr, "\nYear: ", year, "\nTotal acres: ", comma(total.acres), "\nTotal estimated land value: ", dollar(total.value),  "\nLand value per acre: ", dollar(land.value.acre), "\nLand value per acre indexed to 2001 levels: ", percent(land.value.acre.index), sep = ""))) +
  labs(x="", y = "", color="", title = "Land value per farm acre")+
  scale_y_continuous(labels=scales::dollar)+
  coord_cartesian(ylim = seq(-3250,13000, 2500)) +
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_manual(values = color.edr,
                     guide = guide_legend(ncol = 4)) +
  theme(legend.position = "bottom",
        text = element_text(size = 10),
        legend.text = element_text(size = 6),
        axis.text.x = element_text(size = 6))

net.income.acre.edr.plot.legend <- get_legend(net.income.acre.edr.plot + theme(legend.box.margin = margin(0,0,0,0)))

net.income.acre.edr.plot.legend.bottom <- plot_grid(NULL, net.income.acre.edr.plot.legend, NULL, ncol = 3, rel_widths = c(0,.5,3))

girafe(ggobj = plot_grid(net.income.acre.edr.plot + theme(legend.position = "none"), land.value.acre.edr.plot + theme(legend.position = "none"), NULL, net.income.acre.edr.plot.legend.bottom, ncol = 2, rel_heights = c(1, .1)), height_svg = 9) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```

<br>

## Changes in net income vs. changes in land values since year of peak net incomes {.tabset}

In the charts below, land value per acre represents agricultural 2a land. 

The charts provide a comparison of the change in net farm income since the year when net incomes were the highest in each region (which is typically between 2012 and 2014) and the change in agricultural land values since that same year. With the downturn in crop commodities, we would expect land values to be below what they were when net farm income peaked. However, as the previous charts have shown, that may not be the case. 

Interestingly, agricultural land values are higher in 2018 then they were when net farm incomes peaked in nearly all regions across the state. The only areas where this doesn't seem to be the case are in our most rural areas of the state. At the EDR level, 7E - East Central, 6W - Upper Minnesota Valley, and 8 - Southwest all have lower land values in 2018 than when net farm incomes peaked.

Everywhere else, agricultural land values are typically 15% to 40% higher than when net farm incomes peaked.

<br>

```{r prep year of peak net income, include=FALSE, cache=TRUE}
net.inc.peak.year.ruca <- net.income.acre.ruca %>%
  group_by(Dem_Desc) %>%
  top_n(n = 1, wt = net.income.acre) %>%
  ungroup() %>%
  rename(peak.year = 2,
         net.income.peak = 3,
         total.acre.peak = 4,
         net.income.acre.peak = 5)

land.value.peak.year.ruca <- land.value.acre.ruca %>%
  right_join(net.inc.peak.year.ruca[,c(1,2)], by = c("Dem_Desc", "year" = "peak.year")) %>%
  rename(peak.year = 2,
         total.acres.peak = 3,
         total.value.peak = 4,
         land.value.acre.peak = 5)


net.inc.change.peak.year.ruca <- net.income.acre.ruca %>%
  filter(year == 2018) %>%
  left_join(net.inc.peak.year.ruca[,c(1,2,3,4,5)], by = "Dem_Desc") %>%
  mutate(pct.change.net.income.acre = (net.income.acre - net.income.acre.peak) / net.income.acre.peak)

land.value.change.peak.year.ruca <- land.value.acre.ruca %>%
  filter(year == 2018) %>%
  left_join(land.value.peak.year.ruca[,c(1,2,3,4,5)], by = "Dem_Desc") %>%
  mutate(pct.change.land.value.acre = (land.value.acre - land.value.acre.peak) / land.value.acre.peak)

net.inc.land.value.change.peak.year.ruca <- net.inc.change.peak.year.ruca %>%
  left_join(land.value.change.peak.year.ruca[,c(1,4,5,9,10,11)], by = "Dem_Desc") %>%
  gather(key = "income.land", value = "pct.change", 11, 16) %>%
  mutate(income.land = ifelse(income.land == "pct.change.net.income.acre", "Change in net farm income per acre", "Change in land value per farm acre"))


net.inc.peak.year.pr <- net.income.acre.pr %>%
  group_by(planning.region) %>%
  top_n(n = 1, wt = net.income.acre) %>%
  ungroup() %>%
  rename(peak.year = 2,
         net.income.peak = 3,
         total.acre.peak = 4,
         net.income.acre.peak = 5)

land.value.peak.year.pr <- land.value.acre.pr %>%
  right_join(net.inc.peak.year.pr[,c(1,2)], by = c("planning.region", "year" = "peak.year")) %>%
  rename(peak.year = 2,
         total.acres.peak = 3,
         total.value.peak = 4,
         land.value.acre.peak = 5)


net.inc.change.peak.year.pr <- net.income.acre.pr %>%
  filter(year == 2018) %>%
  left_join(net.inc.peak.year.pr[,c(1,2,3,4,5)], by = "planning.region") %>%
  mutate(pct.change.net.income.acre = (net.income.acre - net.income.acre.peak) / net.income.acre.peak)

land.value.change.peak.year.pr<- land.value.acre.pr %>%
  filter(year == 2018) %>%
  left_join(land.value.peak.year.pr[,c(1,2,3,4,5)], by = "planning.region") %>%
  mutate(pct.change.land.value.acre = (land.value.acre - land.value.acre.peak) / land.value.acre.peak)

net.inc.land.value.change.peak.year.pr <- net.inc.change.peak.year.pr %>%
  left_join(land.value.change.peak.year.pr[,c(1,4,5,9,10,11)], by = "planning.region") %>%
  gather(key = "income.land", value = "pct.change", 11, 16) %>%
  mutate(income.land = ifelse(income.land == "pct.change.net.income.acre", "Change in net farm income per acre", "Change in land value per farm acre"),
         planning.region = str_replace(planning.region, "Seven County Mpls-St Paul", "Seven County\nMpls-St Paul"))



net.inc.peak.year.edr <- net.income.acre.edr %>%
  group_by(edr, planning.region) %>%
  top_n(n = 1, wt = net.income.acre) %>%
  ungroup() %>%
  rename(peak.year = 3,
         net.income.peak = 4,
         total.acre.peak = 5,
         net.income.acre.peak = 6)

land.value.peak.year.edr <- land.value.acre.edr %>%
  right_join(net.inc.peak.year.edr[,c(1,2,3)], by = c("edr", "planning.region", "year" = "peak.year")) %>%
  rename(peak.year = 3,
         total.acres.peak = 4,
         total.value.peak = 5,
         land.value.acre.peak = 6)

net.inc.change.peak.year.edr <- net.income.acre.edr %>%
  filter(year == 2018) %>%
  left_join(net.inc.peak.year.edr[,c(1,2,3,4,5,6)], by = c("planning.region", "edr")) %>%
  mutate(pct.change.net.income.acre = (net.income.acre - net.income.acre.peak) / net.income.acre.peak)

land.value.change.peak.year.edr<- land.value.acre.edr %>%
  filter(year == 2018) %>%
  left_join(land.value.peak.year.edr[,c(1,2,3,4,5,6)], by = c("planning.region", "edr")) %>%
  mutate(pct.change.land.value.acre = (land.value.acre - land.value.acre.peak) / land.value.acre.peak)

net.inc.land.value.change.peak.year.edr <- net.inc.change.peak.year.edr %>%
  left_join(land.value.change.peak.year.edr[,c(1,2,5, 6,10,11,12)], by = c("edr", "planning.region")) %>%
  gather(key = "income.land", value = "pct.change", 12, 17) %>%
  mutate(income.land = ifelse(income.land == "pct.change.net.income.acre", "Change in net farm income per acre", "Change in land value per farm acre"))

```

### RUCA

Net farm income peaked in 2013 in entirely rural counties and since that time net income has decreased by 80%, while land values have only decreased 3.4%. 

The rest of the county groups had their net farm income per acre peak in 2012 and since that time have had their net incomes drop a similar amount to entirely rural counties while land value per acre is 25% to 30% higher in 2018 than in 2012. 

<br>

```{r chart change in net inc and land value since peak year rucA}

net.inc.land.value.change.peak.year.ruca.plot <- ggplot(net.inc.land.value.change.peak.year.ruca, aes(x = Dem_Desc, y = pct.change, color = income.land)) +
  geom_hline(yintercept = 0, color = "black") +
  geom_point_interactive(size = 3,aes(data_id = Dem_Desc, tooltip = paste(Dem_Desc, "\nYear of peak net farm income: ", peak.year, "\nNet income per acre during peak year: ", dollar(net.income.acre.peak), "\nLand value per acre during peak year: ", dollar(land.value.acre.peak), "\nNet income per acre in 2018: ", dollar(net.income.acre), "\nLand value per acre in 2018: ", dollar(land.value.acre), "\n", income.land, ": ", percent(pct.change), sep = ""))) +
  labs(x="", y = "", title = "Percent change in net farm income and land value per acre\nsince net farm income peaked")+
        scale_y_continuous(labels=scales::percent)+
        theme_bar+
        theme(legend.position = "bottom",
              legend.title = element_blank())+
        scale_color_manual(values = c("#8c510a", "#076324"))


girafe(ggobj = net.inc.land.value.change.peak.year.ruca.plot, width_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```

### Planning Region

Overall, net farm income per acre peaked between 2012 - 2014 for all planning regions and have decreased between 70% and 95%. Across most regions, land values per acre are still 2% to 40% higher than they were during the peak farm income years. Land values have decreased, however, in Southwest Minnesota where they are 8% of the value they were in 2013, the peak farm income year.


<br>

```{r chart change in net inc and land value since peak year pr}

net.inc.land.value.change.peak.year.pr.plot <- ggplot(net.inc.land.value.change.peak.year.pr, aes(x = planning.region, y = pct.change, color = income.land)) +
  geom_hline(yintercept = 0, color = "black") +
  geom_point_interactive(size = 3,aes(data_id = planning.region, tooltip = paste(planning.region, "\nYear of peak net farm income: ", peak.year, "\nNet income per acre during peak year: ", dollar(net.income.acre.peak), "\nLand value per acre during peak year: ", dollar(land.value.acre.peak), "\nNet income per acre in 2018: ", dollar(net.income.acre), "\nLand value per acre in 2018: ", dollar(land.value.acre), "\n", income.land, ": ", percent(pct.change), sep = ""))) +
  labs(x="", y = "", title = "Percent change in net farm income and land value per acre\nsince net farm income peaked")+
        scale_y_continuous(labels=scales::percent)+
        theme_bar+
        theme(legend.position = "bottom",
              legend.title = element_blank())+
        scale_color_manual(values = c("#8c510a", "#076324"))


girafe(ggobj = net.inc.land.value.change.peak.year.pr.plot, width_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```

### EDR

Except for EDRs 7E - East Central, 6W - Upper Minnesota Valley, and 8 - Southwest, all other EDRs have slightly higher land values in 2018 than they did when net farm incomes peaked. 

<br>

```{r chart change in net inc and land value since peak year edr, fig.height=10}

net.inc.land.value.change.peak.year.edr.plot <- ggplot(net.inc.land.value.change.peak.year.edr, aes(x = edr, y = pct.change, color = income.land)) +
  geom_hline(yintercept = 0, color = "black") +
  facet_wrap(~planning.region, ncol = 2, scales = "free_x") +
  geom_point_interactive(size = 3,aes(data_id = edr, tooltip = paste(edr, "\nYear of peak net farm income: ", peak.year, "\nNet income per acre during peak year: ", dollar(net.income.acre.peak), "\nLand value per acre during peak year: ", dollar(land.value.acre.peak), "\nNet income per acre in 2018: ", dollar(net.income.acre), "\nLand value per acre in 2018: ", dollar(land.value.acre), "\n", income.land, ": ", percent(pct.change), sep = ""))) +
  labs(x="", y = "", title = "Percent change in net farm income and land value per acre\nsince net farm income peaked")+
        scale_y_continuous(labels=scales::percent)+
        theme_bar+
        theme(legend.position = "bottom",
              legend.title = element_blank(),
              text = element_text(size = 10),
              axis.text.x = element_text(size = 6, angle = 15, vjust = .8))+
        scale_color_manual(values = c("#8c510a", "#076324"))

girafe(ggobj = net.inc.land.value.change.peak.year.edr.plot, height_svg = 9) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```




