---
title: "Analyzing industries in economic analysis study"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
runtime: shiny
resource_files:
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.cpg
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.dbf
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.prj
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbn
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbx
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shp.xml
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shx
---

<link rel="stylesheet" href="styles.css" type="text/css">


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
```


```{r loading jon docs and shapefiles}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", size = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.title = element_blank(),
        text = element_text(size = 15),
        plot.caption = element_text(hjust = 0.5, face = "italic"),
        legend.text = element_text(margin = margin(l = 2, r = 5)))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", size = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        text = element_text(size = 15),
        plot.caption = element_text(hjust = 0.5, face = "italic"),
        legend.text = element_text(margin = margin(l = 2, r = 5)))

theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        plot.title = element_text(hjust = 0.5),
        legend.title = element_blank(),
        text = element_text(size = 15),
        plot.caption = element_text(hjust = 0.5, face = "italic"),
        legend.text = element_text(margin = margin(l = 2, r = 5)))

counties.regions <- read_csv("Data/Join docs/county_regions.csv") %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         edr = str_replace(edr, "  ", " "))

regions = counties.regions %>%
  select(5,6) %>%
  distinct(edr, .keep_all = TRUE)

color.ruca <- c("Entirely rural" = "#5CA81F", "Town/rural mix" = "#C7EF99", "Urban/town/rural mix" = "#d8b365", "Entirely urban" = "#a6611a", "Minnesota" = "black")

color.pr <- c("Northwest" = "#810f7c","Northeast" = "#fe9929", "Central" = "#076324", "Seven County Mpls-St Paul" = "#d8b365", "Southwest" = "#1f78b4", "Southeast" = "#d7301f", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.counties <- scale_color_brewer(palette = "Dark2",
                       guide = guide_legend(ncol = 3))

mn_counties <- st_read("Data/Shapefiles/county shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE)
```

```{r master objects, include=FALSE, cache=TRUE}
master.qcew.pr <- read_csv("Data/Lazarus/lazarus-qcew-pr.csv") %>%
  rename(planning.region = areaname) %>%
  mutate(planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = str_replace(planning.region, ", MN", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"))

master.qcew.edr <- read_csv("Data/Lazarus/lazarus-qcew-edr.csv") %>%
  mutate(areaname = str_replace(areaname, "  ", " ")) %>%
  rename("edr" = "areaname") %>%
  left_join(regions, by = "edr") %>%
  mutate(planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

master.qcew.county <- read_csv("Data/Lazarus/lazarus-qcew-county.csv") %>%
  mutate(areaname = str_replace(areaname, " County", ""),
         areaname = str_replace(areaname, "Saint Louis", "St. Louis")) %>%
  left_join(counties.regions[,c(1,2)], by = c("areaname" = "Name"))
```

<br>

# Overview of the report

Bill Lazarus, an Applied Economics professor at the University of Minnesota, provided a report analyzing the impacts of reduced farm spending in regions across the state. You can view the full report [here](https://www.dropbox.com/home/CRPD%20Team%20Folder/Research%202020/Ag%20and%20rural%20economy/Literature%20Review?preview=Impacts+of+Reduced+Farm+Spending+on+Rural+Minnesota+Communities103019.pdf).

Due to lower crop prices, crop producers are being encouraged to "tighten their belts", resulting in less spending for various products and equipment. Therefore, this report analyzed the impacts on communities with reduced spending in four main categories;

1. farm machinery,
2. building purchases,
3. cropland rental rates, and
4. fertilizer.

It's important to note that farm household discretionary spending on goods such as televisions and sofas might also see cutbacks with reduced incomes, but is NOT considered in this analysis. However, there are so few households farming presently that it isn't considered to be a major source of trouble.

[the following is taken directly from the executive summary of Dr. Lazarus' report]

Results show that job losses at farm machinery dealers, construction firms, fertilizer dealers, and in other sectors supplying them were 6,480 statewide, or 15,720 with secondary ripples. These job losses would be a severe economic hit to those workers and their families. In percentage terms the total impact on the state’s economy with secondary ripple effects is 0.47%.

Another measure of community impact is that labor income would decline by $193/person statewide (based on 3.7 million total employment). Figure 7 shows that the industry sectors that are most affected are trade and the financial and real estate sector.

Farm real estate value changes across the regions appear to be consistent with changes in land rents except that values have held up better in the southeastern region compared to rents there.

# Summary

When we dig deeper into smaller geographic areas in the state we can begin to see that there is more fluctuation and likely areas being more impacted than what the Lazarus study shows. In particular, EDRs 6W - Upper Minnesota Valley and EDR 8 - Southwest in Southwest Minnesota seem to have experienced significant decreases in their employment over the last few years within the industries identified in the Lazarus study.

However, the overall extent of employment in these industries even at the county level barely exceed 8% at their highest levels, and are below 2% across a large majority of counties. But, we can identify a couple of characteristics of which counties are most likely to be impacted;

* have had 3% or more of their employment in industries identified in the Lazarus study at some point since 2001; and,
* are located in Southwest Minnesota and have 3% or more of the employment in industries identified in the Lazarus study at some point since 2001.

When analyzing counties that fit these characteristics, the trends show a bit of a mixed bag with some counties experiencing decreases in these industries while other counties experiencing increases.

Counties experiencing a decrease in employment in these industries;

* Grant
* Stevens
* McLeod
* Swift
* Chippewa
* Pipestone
* Rock
* Martin
* Faribault

Counties experiencing an increase in employment in these industries;

* Marshall
* Wilkin
* Meeker
* Renville
* Lac qui Parle
* Redwood
* Watonwan

Due to such low numbers of employment, the data is a bit spotty and only a few industries actually show up at the county level when we try to examine the trends of individual industries. This, in itself, is indicative of the lack of impact these industries really have on overall employment in Minnesota.

However, overall there seems to be two main industries that indicate whether a county is increasing or decreasing in employment in these ag-related industries as a whole;

farm product raw material merchant wholesalers, and
farm supplies merchant wholesalers.

If these industries were increasing in the county, than it was likely that the county was experiencing an increase in the ag-related industries as a whole, and vice-versa.

# Analysis

The report produced by Bill Lazarus provides a more holistic analysis of the impacts that low crop commodity prices over a long duration of time has had in farmer spending in specific industries related to crop production. The analysis used IMPLAN and data from sources that are not publicly available. The one limit to this study is that, due to data limitations, analysis was only to go down as far as the planning region of each state.

However, the report did reveal specific industries that should be, theoretically, impacted the most. These industries are;

* Financial, Insurance, Real Estate
* Trade
* Construction
* Manufacturing
* Transport, Utilities, Government & Other

At a regional level, Lazarus's report showed that the impacts from lower farmer spending in industries related to crop production to be relatively minor. However, it could be that the impacts are diluted at the planning region level, and there may be some counties where these industries are concentrated that are especially feel the affects. This report seeks to identify those counties, if any. 

The analysis will assess employment and wage data for the industries identified in the Lazarus report. Below is a table providing the industry category that was provided in the Lazarus report with the coinciding NAIC industry category that the data in our report was pulled.

<br>

```{r lazarus category table}

lazarus.cat <- read_csv("Data/Lazarus/Lazarus industry categories.csv")

kable(format = "html", lazarus.cat, escape = F) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

```

<br>

## Impacted ag industries as a percent of total employment {.tabset}

Similar to the Lazarus study, the percentage of employment in industries identified as being most impacted by crop commodity prices has stayed consistent across all of Minnesota. The highest percentage is in SW Minnesota.

However, as we look at the percentages in smaller geographic areas we can begin to see that there is more fluctuation and likely areas being more impacted. In particular, EDRs 6W - Upper Minnesota Valley and EDR 8 - Southwest in Southwest Minnesota seem to have experienced significant decreases in their employment over the last few years. 

The map highlights that employment in these industries are concentrated in counties located in Southwest Minnesota and to some extent, up along the South and North Dakota borders.

```{r prep combined ag industries as pct of total, include=FALSE, cache=TRUE}

emp.pr <- master.qcew.pr %>%
  select(1,2,5,28) %>%
  group_by(periodyear, planning.region) %>%
  summarise(ag.emp = sum(empYear[naicstitle != "Total, All Industries"], na.rm = TRUE),
            total.emp = sum(empYear[naicstitle == "Total, All Industries"], na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(ag.pct.total = ag.emp / total.emp)

emp.edr <- master.qcew.edr %>%
  select(1,2,5,28, 83) %>%
  group_by(periodyear, edr, planning.region) %>%
  summarise(ag.emp = sum(empYear[naicstitle != "Total, All Industries"], na.rm = TRUE),
            total.emp = sum(empYear[naicstitle == "Total, All Industries"], na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(ag.pct.total = ag.emp / total.emp)

emp.county <- master.qcew.county %>%
  select(1,2,5,28,83) %>%
  group_by(periodyear, areaname, countyfp) %>%
  summarise(ag.emp = sum(empYear[naicstitle != "Total, All Industries"], na.rm = TRUE),
            total.emp = sum(empYear[naicstitle == "Total, All Industries"], na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(ag.pct.total = ag.emp / total.emp,
         ag.pct.total.bins = cut(ag.pct.total,
                                 breaks = c(-1, .01, .02, .03, .04, 1),
                                 labels = c("0% to 1%", "1% to 2%", "2% to 3%", "3% to 4%", "More than 4%"))) %>%
  left_join(mn_counties[,c(4,7)], by = c("countyfp" = "FIPS_CODE")) 
```

<br>

### Planning Region

The region with the highest percentage of their employment in industries identified as being most impacted by the reduction in farmer spending is Southwest Minnesota with 3%. 

<br>

```{r chart ag emp as pct total emp pr}
emp.pr.plot <- ggplot(emp.pr, aes(periodyear, ag.pct.total, color = planning.region)) +
  geom_smooth(se = FALSE, size = 2) +
  geom_point_interactive(size = 3, aes(data_id = ag.pct.total, tooltip = paste(planning.region, "\nYear: ", periodyear, "\nEmployment in ag industries: ", comma(ag.emp), "\nTotal employment: ", comma(total.emp), "\nAg employment as % of total employment: ", percent(ag.pct.total), sep = ""))) +
  labs(x="", y = "", color="", title = "Lazarus ag employment as a % of total employment")+
  scale_y_continuous(labels=scales::percent)+
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_manual(values= color.pr,
                     guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom")

girafe(ggobj = emp.pr.plot, width_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```

### EDR

The edr's with the highest percentage of employment in industries identified as being most impacted in the Lazarus study are located in the Northwest and Southwest regions of the state. In particular, EDR 1 - Northwest, EDR 4 - West Central, EDR 6W - Southwest, and EDR 8 - Southwest. 

The regions that seem to be most impacted by the downturn in the prices of crop commodities seem to be in EDR 6W and 8 where they used to have 6% or more of their employment based in these industries and have fallen to between 2% and 4%.

<br>

```{r chart ag emp as pct total emp edr, fig.height=9}
emp.edr.plot <- ggplot(emp.edr, aes(periodyear, ag.pct.total, color = edr)) +
  facet_wrap(~planning.region, ncol = 2) +
  geom_smooth(se = FALSE, size = 2) +
  geom_point_interactive(size = 3, aes(data_id = ag.pct.total, tooltip = paste(edr, "\nYear: ", periodyear, "\nEmployment in ag industries: ", comma(ag.emp), "\nTotal employment: ", comma(total.emp), "\nAg employment as % of total employment: ", percent(ag.pct.total), sep = ""))) +
  labs(x="", y = "", color="", title = "Lazarus ag employment as a % of total employment")+
  scale_y_continuous(labels=scales::percent)+
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_manual(values= color.edr,
                     guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(size = 8),
        legend.text = element_text(size = 8),
        text = element_text(size = 12))

girafe(ggobj = emp.edr.plot, height_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```

### 2018 county map

It's pretty easy to see that the western side of the state has the highest concentration of industries identified in Lazarus's report. The Southwest region, in particular, has many counties with 2% or more of their employment in these industries. Lac qui Parle, Yellow Medicine and Redwood each have employment of 4% or more in these industries.

<br>

```{r map ag employment as pct of total in 2018}
emp.county.map <- ggplot(filter(emp.county, periodyear == 2018)) +
  geom_sf_interactive(aes(fill = ag.pct.total.bins, tooltip = paste(areaname, "\nEmployment in ag industries: ", comma(ag.emp), "\nTotal employment: ", comma(total.emp), "\nAg employment as % of total employment: ", percent(ag.pct.total), sep = ""), data_id = areaname, geometry = geometry)) +
  scale_fill_manual(values = c("white", "#edf8fb", "#b2e2e2", "#2ca25f", "#006d2c"))+
  theme_sf+
  labs(title="2018 Lazarus ag industries employment as % of total employment") +
  theme(text = element_text(size = 10))

girafe(ggobj = emp.county.map, width_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))


```

<br>

## Counties most impacted by the industries identified in Lazarus study {.tabset}

The charts above identify a couple of characteristics of which counties are most likely to be impacted;

* have had 3% or more of their employment in industries identified in the Lazarus study at some point since 2001; and,
* are located in Southwest Minnesota and have 3% or more of the employment in industries identified in the Lazarus study at some point since 2001.

There were 22 counties that fit one of the criteria above and the  charts below provide the employment trends within the specific industries identified in the Lazarus study.

The trends show a bit of a mixed bag with some counties experiencing decreases in these industries while other counties experiencing increases. 

Counties experiencing a decrease in employment in these industries;

* Grant
* Stevens
* McLeod
* Swift
* Chippewa
* Pipestone
* Rock
* Martin
* Faribault

Counties experiencing an increase in employment in these industries;

* Marshall
* Wilkin
* Meeker
* Renville
* Lac qui Parle
* Redwood
* Watonwan

<br>

```{r prep filtered counties, include=FALSE, cache=TRUE}
filt.counties <- emp.county %>%
  filter(ag.pct.total > .0299999) %>%
  distinct(areaname, .keep_all = TRUE)

filt.trends.counties <- emp.county %>%
  right_join(filt.counties[,c(3)], by = "countyfp") %>%
  group_by(areaname, countyfp) %>%
  mutate(ag.emp.change = ag.emp / ag.emp[periodyear==2001]) %>%
  left_join(counties.regions[,c(1,4,5,6)], by = "countyfp")
```

### Northwest

Employment in the specific ag industries related to the Lazarus study in each EDR in Northwest Minnesota is below 150. However, there are a couple of counties here where the downturn in crop commodity prices may be impacting employment to a significant degree. Grant and Stevens Counties in EDR 4 - West Central each have a fairly high percentage of their employment in these specific industries (between 3% and 7%) and seem to be experiencing declines in these industries over the last few years.

Pope looks like it might be bad but it's such a small percentage of total employment that the impacts are likely very limited.

On the other hand, Marshall and Wilkin Counties have experienced some healthy growth in these industries over the last few years. 

<br>

```{r charts trends of filtered counties nw}
filt.trends.counties.nw.plot <- ggplot(filter(filt.trends.counties, planning.region == "Northwest"), aes(periodyear, ag.emp, color = areaname)) +
  facet_wrap(~edr, ncol = 2) +
  geom_smooth(se = FALSE, size = 2) +
  geom_point_interactive(size = 3, aes(data_id = ag.emp, tooltip = paste(areaname, "\nYear: ", periodyear, "\nEmployment in ag industries: ", comma(ag.emp), "\nTotal employment: ", comma(total.emp), "\nAg employment as % of total employment: ", percent(ag.pct.total), sep = ""))) +
  labs(x="", y = "", color="", title = "Ag employment as percent of total employment")+
  scale_y_continuous(labels=scales::comma)+
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_discrete(guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(size = 10))

girafe(ggobj = filt.trends.counties.nw.plot, width_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```

### Central

For the most part, employment in these specific industries seems to be increasing in EDR 6E. McLeod took a significant dip in 2017 which is likely due to a plan closing in that EDR. It might be worth checking to see what happened there.

<br>

```{r charts trends of filtered counties central}
filt.trends.counties.c.plot <- ggplot(filter(filt.trends.counties, planning.region == "Central"), aes(periodyear, ag.emp, color = areaname)) +
  facet_wrap(~edr, ncol = 2) +
  geom_smooth(se = FALSE, size = 2) +
  geom_point_interactive(size = 3, aes(data_id = ag.emp, tooltip = paste(areaname, "\nYear: ", periodyear, "\nEmployment in ag industries: ", comma(ag.emp), "\nTotal employment: ", comma(total.emp), "\nAg employment as % of total employment: ", percent(ag.pct.total), sep = ""))) +
  labs(x="", y = "", color="", title = "Ag employment as percent of total employment")+
  scale_y_continuous(labels=scales::comma)+
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_discrete(guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(size = 10))

girafe(ggobj = filt.trends.counties.c.plot, width_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```


### Southwest

In EDR 6W, I would have expected a more prominent trend of EDRs experiencing a decrease in their employment in these specific industries. It turns out that only Swift and Chippewa counties are the only ones experiencing declines. On the flip side, Lac qui Parle is experiencing some healthy increases.

In EDR 8, Pipestone and Rock Counties have experienced some declines while Redwood is experiencing an increase in the last few years.

In EDR 9, Martin County has experienced significant peaks and valleys in their employment in these specific industries. IT peaked in 2001 at 436 (4% of total employment) and has cratered to below 50 in the last few years. Faribault has also experience a steady decline. Watonwan, on the other hand, has experienced an increase over the last few years.

<br>

```{r charts trends of filtered counties sw, fig.height=6}
filt.trends.counties.sw.plot <- ggplot(filter(filt.trends.counties, planning.region == "Southwest"), aes(periodyear, ag.emp, color = areaname)) +
  facet_wrap(~edr, ncol = 2) +
  geom_smooth(se = FALSE, size = 1) +
  geom_point_interactive(size = 2, aes(data_id = ag.emp, tooltip = paste(areaname, "\nYear: ", periodyear, "\nEmployment in ag industries: ", comma(ag.emp), "\nTotal employment: ", comma(total.emp), "\nAg employment as % of total employment: ", percent(ag.pct.total), sep = ""))) +
  labs(x="", y = "", color="", title = "Ag employment as percent of total employment")+
  scale_y_continuous(labels=scales::comma)+
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_discrete(guide = guide_legend(ncol = 4)) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(size = 8),
        text = element_text(size = 10))

girafe(ggobj = filt.trends.counties.sw.plot, height_svg = 5) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```


<br>

## Specific ag industries driving change {.tabset}

The next question to ask is what industries identified in the Lazarus report are driving the increases and decreases in those ag-related employment sectors. Below are the 16 counties identified above experiencing somewhat clear trends of decreasing and increasing employment in the industries identified in the Lazarus study and the employment trends in each of those industries.

Due to such low numbers of employment, the data is a bit spotty and only a few industries actually show up at the county level. This, in itself, is indicative of the lack of impact these industries really have on overall employment in Minnesota. 

Overall, there are two main industries that seem to indicate whether a county is increasing or decreasing in employment in these ag-related industries as a whole;

1. farm product raw material merchant wholesalers, and
2. farm supplies merchant wholesalers.

If these industries were increasing in the county, than it was likely that the county was experiencing an increase in the ag-related industries as a whole, and vice-versa.

<br>

```{r prep industry trends for impacted counties, include=FALSE, cache = TRUE}
impacted.counties <- master.qcew.county %>%
  filter(areaname %in% c("Grant", "Stevens", "McLeod", "Swift", "Chippewa", "Pipestone", "Rock", "Martin", "Faribault", "Marshall", "Wilkin", "Meeker", "Renville", "Lac qui Parle", "Redwood", "Watonwan"),
         naicstitle != "Total, All Industries") %>%
  select(1,2,5,28,83) %>%
  group_by(areaname, countyfp) %>%
  complete(periodyear, naicstitle, fill = list(c(empYear = 0))) %>%
  mutate(empYear = ifelse(is.na(empYear), 0, empYear),
         inc.dec = ifelse(areaname %in% c("Grant", "Stevens", "McLeod", "Swift", "Chippewa", "Pipestone", "Rock", "Martin", "Faribault"), "Decreasing", "Increasing"),
         data_id = seq(n())) %>%
  left_join(counties.regions[,c(1,5,6)], by = "countyfp")

```

### Counties with increasing employment in Lazarus industries{.tabset}

In counties where there is increasing employment in industries identified in the Lazarus study, there seems to be two industries within the retail and trade sector driving the increases;

1. Farm product raw material merchant wholesales, and
2. Farm supplies merchant wholesalers.

<br>

```{r charts lazarus industry trends increasing counties, fig.height=9}

impacted.counties.plot <- ggplot(filter(impacted.counties, inc.dec == "Increasing"), aes(periodyear, empYear, color = naicstitle)) +
  facet_wrap(~areaname, ncol = 2) +
  geom_point_interactive(size = 2, aes(data_id = data_id, tooltip = paste(areaname, "Year: ", periodyear, "\nNAICS title: ", naicstitle, "\nEmployment: ", comma(empYear), sep = ""))) +
  labs(x="", y = "", color="", title = "Employment by industry and county")+
  scale_y_continuous(labels=scales::comma)+
    scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_discrete(guide = guide_legend(ncol = 2)) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        legend.text = element_text(size = 8),
        text = element_text(size = 10))

girafe(ggobj = impacted.counties.plot, height_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```


### Counties with decreasing employment in Lazarus industries

The overall industry driver of declines these counties are;

* farm and garden machinery and equipment merchant wholesalers,
* farm product raw material merchant wholesalers
* farm supplies merchant wholesalers.

Below is a list of each of the counties and the general trends of their industries.

Chippewa: Decreases in farm and garden machinery and equipment merchant wholesalers and farm supplies merchant wholesalers.

Faribault: Declines in Farm product raw material merchant wholesalers, and farm and garden machinery and equipment merchant wholesalers.

Grant: Declines in farm product raw material merchant wholesalers and the ups and downs of farm supplies merchant wholesalers.

Martin: Significant appearances and disappearances of farm supplies merchant wholesalers and farm product raw material merchant wholesalers.

McLeod: Disappearance of farm and garden machinery and equipment merchant wholesalers, and ag implement manufacturing. But seeing increases in farm supplies merchant wholesalers and commercial and institutional building construction.

Pipestone: Decline in farm product raw materials merchant wholesalers and disappearance of farm and garden machinery and equipment merchant wholesalers.

Rock: Disappearance of farm and garden machinery and equipment merchant wholesalers and farm supplies merchant wholesaler.

Stevens: Decline in farm product raw materials merchant wholesalers with some ups and downs in the farm and garden machinery and equipment merchant wholesalers. Increases in farm supplies merchant wholesalers.

Swift: Decreases in farm product raw material merchant wholesalers.


<br>

```{r charts lazarus industry trends decreasing counties, fig.height=9}

impacted.counties.decreasing.plot <- ggplot(filter(impacted.counties, inc.dec == "Decreasing"), aes(periodyear, empYear, color = naicstitle)) +
  facet_wrap(~areaname, ncol = 2) +
  geom_point_interactive(size = 2, aes(data_id = data_id, tooltip = paste(areaname, "Year: ", periodyear, "\nNAICS title: ", naicstitle, "\nEmployment: ", comma(empYear), sep = ""))) +
  labs(x="", y = "", color="", title = "Employment by industry and county")+
  scale_y_continuous(labels=scales::comma)+
    scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_discrete(guide = guide_legend(ncol = 2)) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        legend.text = element_text(size = 8),
        text = element_text(size = 10))

girafe(ggobj = impacted.counties.decreasing.plot, height_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```


