---
title: "Workspace"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
runtime: shiny
resource_files:
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.cpg
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.dbf
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.prj
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbn
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbx
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shp.xml
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shx
---

<link rel="stylesheet" href="styles.css" type="text/css">


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
```

```{r loading jon docs and shapefiles, cache=TRUE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", size = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.title = element_blank(),
        text = element_text(size = 15),
        plot.caption = element_text(hjust = 0.5, face = "italic"),
        legend.text = element_text(margin = margin(l = 2, r = 5)))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", size = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        text = element_text(size = 15),
        plot.caption = element_text(hjust = 0.5, face = "italic"),
        legend.text = element_text(margin = margin(l = 2, r = 5)))

theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        plot.title = element_text(hjust = 0.5),
        legend.title = element_blank(),
        text = element_text(size = 15),
        plot.caption = element_text(hjust = 0.5, face = "italic"),
        legend.text = element_text(margin = margin(l = 2, r = 5)))

counties.regions <- read_csv("Data/Join docs/county_regions.csv") %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc))

regions = counties.regions %>%
  select(5,6) %>%
  distinct(edr, .keep_all = TRUE)

edr.pr <- counties.regions %>%
  distinct(edr, .keep_all = TRUE) %>%
  select(5,6) %>%
  mutate(edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""))

color.ruca <- c("Entirely rural" = "#5CA81F", "Town/rural mix" = "#C7EF99", "Urban/town/rural mix" = "#d8b365", "Entirely urban" = "#a6611a", "Minnesota" = "black")

color.pr <- c("Northwest" = "#810f7c","Northeast" = "#fe9929", "Central" = "#076324", "Seven County Mpls-St Paul" = "#d8b365", "Southwest" = "#1f78b4", "Southeast" = "#d7301f", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.counties <- scale_color_brewer(palette = "Dark2",
                       guide = guide_legend(ncol = 3))

color.ag.ind <- c("Crop Production" = "#a6cee3", "Animal Production and Aquaculture" = "#1f78b4", "Support Activities for Crop Production" = "#b2df8a", "Support Activities for Animal Production" = "#33a02c", "Food Manufacturing" = "#fb9a99", "Agricultural Implement Manufacturing" = "#e31a1c", "Farm Product Raw Material Merchant Wholesalers" = "#fdbf6f", "Farm Supplies Merchant Wholesalers" = "#ff7f00", "Pesticide, Fertilizer, and Other Agricultural Chemical Manufacturing" = "#cab2d6", "Total, All Industries" = "black", "NA" = "white")

mn_counties <- st_read("Data/Shapefiles/county shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE)
```

```{r prep master wage file, include = FALSE, cache = TRUE}
ag.ind.wages.county <- read_csv("Data/Earnings/Master-ag-related-ind-wages-counties.csv")

ag.ind.wages.pr <- read_csv("Data/Earnings/Master-ag-related-ind-wages-pr.csv") %>%
  mutate(planning.region = str_replace(planning.region, " Minnesota", ""))

ag.ind.wages.edr <- read_csv("Data/Earnings/Master-ag-related-ind-wages-edr.csv") %>%
  left_join(edr.pr, by = "edr") %>%
  mutate(planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"))

```

```{r prep master ag employment objects, include=FALSE, cache=TRUE}
master.ag.emp.county <- read_csv("Data/Employment/Master-agrelated-emp-county.csv")

master.ag.emp.comb.county <- master.ag.emp.county %>%
  group_by(periodyear, Name, countyfp, Dem_Desc, edr, planning.region) %>%
  summarise(ag.emp = sum(emp[naicstitle != "Total, All Industries"]),
            total.emp = sum(emp[naicstitle == "Total, All Industries"])) %>%
  ungroup()

master.ag.emp.ruca <- read_csv("Data/Employment/Master-agrelated-emp-ruca.csv") %>%
  mutate(Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

master.ag.emp.comb.ruca <- master.ag.emp.ruca %>%
  group_by(periodyear, Dem_Desc) %>%
  summarise(ag.emp = sum(emp[naicstitle != "Total, All Industries"]),
            total.emp = sum(emp[naicstitle == "Total, All Industries"])) %>%
  ungroup()

master.ag.emp.pr <- read_csv("Data/Employment/Master-agrelated-emp-pr.csv") %>%
  mutate(planning.region = str_replace(planning.region, ", MN", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"))

master.ag.emp.comb.pr <- master.ag.emp.pr %>%
  group_by(periodyear, planning.region) %>%
  summarise(ag.emp = sum(emp[naicstitle != "Total, All Industries"]),
            total.emp = sum(emp[naicstitle == "Total, All Industries"])) %>%
  ungroup()


master.ag.emp.edr <- read_csv("Data/Employment/Master-agrelated-emp-edr.csv") %>%
  mutate(planning.region = str_replace(planning.region, ", MN", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

master.ag.emp.comb.edr <- master.ag.emp.edr %>%
  group_by(periodyear, edr, planning.region) %>%
  summarise(ag.emp = sum(emp[naicstitle != "Total, All Industries"]),
            total.emp = sum(emp[naicstitle == "Total, All Industries"])) %>%
  ungroup()

```

```{r master combined wages and employment, include=FALSE, cache=TRUE}
ag.wages.emp.county <- ag.ind.wages.county %>%
  left_join(master.ag.emp.county[,c(1,2,5,6)], by = c("countyfp", "periodyear", "naicstitle")) %>%
  mutate(wages.emp = wages / emp)

ag.wages.emp.ruca <- ag.wages.emp.county %>%
  group_by(periodyear, Dem_Desc, naicstitle) %>%
  summarise(wages = sum(wages, na.rm = TRUE),
            emp = sum(emp, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(wages.emp = wages / emp,
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

ag.wages.emp.pr <- ag.ind.wages.pr %>%
  left_join(master.ag.emp.pr[,c(1,2,3,5)], by = c("planning.region", "periodyear", "naicstitle")) %>%
  mutate(wages.emp = wages / emp,
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"))

ag.wages.emp.edr <- ag.ind.wages.edr %>%
  left_join(master.ag.emp.edr[,c(1,2,3,5)], by = c("edr", "periodyear", "naicstitle")) %>%
  mutate(wages.emp = wages / emp,
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))
```

```{r prep ag wages per emp combined, include=FALSE, cache=TRUE}

ag.wages.emp.comb.ruca <- ag.wages.emp.ruca %>%
  group_by(periodyear, Dem_Desc) %>%
  summarise(ag.wages = sum(wages[naicstitle != "Total, All Industries"]),
            total.wages = sum(wages[naicstitle == "Total, All Industries"]),
            ag.emp = sum(emp[naicstitle != "Total, All Industries"]),
            total.emp = sum(emp[naicstitle == "Total, All Industries"])) %>%
  ungroup() %>%
  mutate(ag.wages.per.emp = ag.wages / ag.emp,
         total.wages.per.emp = total.wages / total.emp)

ag.wages.emp.comb.pr <- ag.wages.emp.pr %>%
  group_by(periodyear, planning.region) %>%
  summarise(ag.wages = sum(wages[naicstitle != "Total, All Industries"]),
            total.wages = sum(wages[naicstitle == "Total, All Industries"]),
            ag.emp = sum(emp[naicstitle != "Total, All Industries"]),
            total.emp = sum(emp[naicstitle == "Total, All Industries"])) %>%
  ungroup() %>%
  mutate(ag.wages.per.emp = ag.wages / ag.emp,
         total.wages.per.emp = total.wages / total.emp)

ag.wages.emp.comb.edr <- ag.wages.emp.edr %>%
  group_by(periodyear, edr, planning.region) %>%
  summarise(ag.wages = sum(wages[naicstitle != "Total, All Industries"]),
            total.wages = sum(wages[naicstitle == "Total, All Industries"]),
            ag.emp = sum(emp[naicstitle != "Total, All Industries"]),
            total.emp = sum(emp[naicstitle == "Total, All Industries"])) %>%
  ungroup() %>%
  mutate(ag.wages.per.emp = ag.wages / ag.emp,
         total.wages.per.emp = total.wages / total.emp)

ag.wages.emp.comb.county <- ag.wages.emp.county %>%
  group_by(periodyear, areaname, countyfp) %>%
  summarise(ag.wages = sum(wages[naicstitle != "Total, All Industries"]),
            total.wages = sum(wages[naicstitle == "Total, All Industries"]),
            ag.emp = sum(emp[naicstitle != "Total, All Industries"]),
            total.emp = sum(emp[naicstitle == "Total, All Industries"])) %>%
  ungroup() %>%
  mutate(ag.wages.per.emp = ag.wages / ag.emp,
         total.wages.per.emp = total.wages / total.emp,
         ag.wages.per.emp = ifelse(areaname == "Mahnomen", NA, ag.wages.per.emp)) %>%
  left_join(mn_counties[,c(4,7)], by = c("countyfp" = "FIPS_CODE"))

```

```{r prep each industry wage per employment statewide, include=FALSE, cache=TRUE}

ag.wages.per.emp.ind.mn <- read_csv("Data/Employment/qcew-ag-related-industries-mn.csv") %>%
  select(1,2,5,28,30) %>%
  rename(emp = empYear,
         wages = totwageYear) %>%
  group_by(periodyear, naicstitle) %>%
  summarise(wages = sum(wages, na.rm = TRUE),
            emp = sum(emp, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(wages.per.emp = wages / emp)
```

<br>

## Concentration of ag-related industries {.tabset}

Summary here

<br>

```{r prep concentration of ag-related industries, include=FALSE, cache=TRUE}

ag.ind.conc.ruca <- ag.wages.emp.ruca %>%
  group_by(periodyear, naicstitle) %>%
  mutate(pct.ind.emp = emp / sum(emp)) %>%
  ungroup()

ag.ind.conc.pr <- ag.wages.emp.pr %>%
  group_by(periodyear, naicstitle) %>%
  mutate(pct.ind.emp = emp / sum(emp)) %>%
  ungroup()

ag.ind.conc.edr <- ag.wages.emp.edr %>%
  group_by(periodyear, naicstitle) %>%
  mutate(pct.ind.emp = emp / sum(emp)) %>%
  ungroup()

```

### RUCA

Due to the size of the shear workforce located in our entirely urban counties, they have a majority of the workforce in each of these industries except for animal production and support actitivies for crop production. 

For the most part, wages per employment are similar across all of the regions except for ag impement manufacturing, farm product raw material merchant wholesalers, and farm supplies merchant wholesalers where the values are significantl higher than other county group types. 

One other interesting element is that our entirely rural county group has a very low percentage of the workforce in each of these industries except for support activities for crop production where it has 40% of the total workforce in that industry.

<br>

```{r chart ag ind concentration, fig.height=8}
ag.ind.conc.ruca.plot <- ggplot(filter(ag.ind.conc.ruca, periodyear == 2018, naicstitle != "Total, All Industries"), aes(pct.ind.emp, wages.emp, color = Dem_Desc)) +
  facet_wrap(~naicstitle, ncol = 2) +
  geom_smooth(se = FALSE, size = 2) +
  geom_point_interactive(size = 3, aes(data_id = pct.ind.emp, tooltip = paste(Dem_Desc, "\nNAICS: ", naicstitle, "\nEmployment in industry: ", comma(emp), "\nTotal wages earned in industry: ", dollar(wages), "\nWages per employment: ", dollar(wages.emp), "\nPercent of industry employment located in this region: ", percent(pct.ind.emp), sep = "")), show.legend = FALSE) +
  geom_text_repel(size = 3, aes(label = Dem_Desc), show.legend = FALSE) +
  labs(x="", y = "", color="", title = "Percent of industry employment located in each region by wage per employment earned")+
  scale_y_continuous(labels=scales::dollar)+
  scale_x_continuous(labels=scales::percent) +
  theme_line+
  scale_color_manual(values= color.ruca,
                     guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 12))

girafe(ggobj = ag.ind.conc.ruca.plot, height_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```

### Planning Region

Breaking it down by planning region shows that it isn't, necessarily, the seven-county metro that has a higher percentage of the workforce across all these industries, but rather, there are metropolitan counties in the other regions that also share a considerable amount of the workforce.



<br>

```{r chart ag ind concentration pr, fig.height=8}
ag.ind.conc.pr.plot <- ggplot(filter(ag.ind.conc.pr, periodyear == 2018, naicstitle != "Total, All Industries"), aes(pct.ind.emp, wages.emp, color = planning.region)) +
  facet_wrap(~naicstitle, ncol = 2) +
  geom_smooth(se = FALSE, size = 2) +
  geom_point_interactive(size = 3, aes(data_id = pct.ind.emp, tooltip = paste(planning.region, "\nNAICS: ", naicstitle, "\nEmployment in industry: ", comma(emp), "\nTotal wages earned in industry: ", dollar(wages), "\nWages per employment: ", dollar(wages.emp), "\nPercent of industry employment located in this region: ", percent(pct.ind.emp), sep = "")), show.legend = FALSE) +
  geom_text_repel(size = 3, aes(label = planning.region), show.legend = FALSE) +
  labs(x="", y = "", color="", title = "Percent of industry employment located in each region by wage per employment earned")+
  scale_y_continuous(labels=scales::dollar)+
  scale_x_continuous(labels=scales::percent) +
  theme_line+
  scale_color_manual(values= color.pr,
                     guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 12))

girafe(ggobj = ag.ind.conc.pr.plot, height_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```

### EDR

Analysis here

<br>



