---
title: "Workspace"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
runtime: shiny
resource_files:
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.cpg
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.dbf
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.prj
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbn
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbx
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shp.xml
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
```

```{r loading jon docs and shapefiles, cache=TRUE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", size = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.title = element_blank(),
        text = element_text(size = 15),
        plot.caption = element_text(hjust = 0.5, face = "italic"),
        legend.text = element_text(margin = margin(l = 2, r = 5)))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", size = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        text = element_text(size = 15),
        plot.caption = element_text(hjust = 0.5, face = "italic"),
        legend.text = element_text(margin = margin(l = 2, r = 5)))

theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        plot.title = element_text(hjust = 0.5),
        legend.title = element_blank(),
        text = element_text(size = 15),
        plot.caption = element_text(hjust = 0.5, face = "italic"),
        legend.text = element_text(margin = margin(l = 2, r = 5)))

counties.regions <- read_csv("Data/Join docs/county_regions.csv") %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc))

color.ruca <- c("Entirely rural" = "#5CA81F", "Town/rural mix" = "#C7EF99", "Urban/town/rural mix" = "#d8b365", "Entirely urban" = "#a6611a", "Minnesota" = "black")

color.pr <- c("Northwest" = "#810f7c","Northeast" = "#fe9929", "Central" = "#076324", "Seven County Mpls-St Paul" = "#d8b365", "Southwest" = "#1f78b4", "Southeast" = "#d7301f", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.counties <- scale_color_brewer(palette = "Dark2",
                       guide = guide_legend(ncol = 3))

mn_counties <- st_read("Data/Shapefiles/county shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE)
```

```{r prep master wage file, include = FALSE, cache = TRUE}
ag.ind.wages.county <- read_csv("Data/Earnings/Master-ag-related-ind-wages-counties.csv")

ag.ind.wages.pr <- read_csv("Data/Earnings/Master-ag-related-ind-wages-pr.csv") %>%
  mutate(planning.region = str_replace(planning.region, " Minnesota", ""))

ag.ind.wages.edr <- read_csv("Data/Earnings/Master-ag-related-ind-wages-edr.csv") 

```

<br>

## Change in ag-related wages vs. total wages {.tabset}

The charts below provide the change in ag-related wages and total wages by indexing the values to their 2001 levels. The trend to look for is how the fluctuations in farm earnings “appear” in the total earnings chart. If total earnings show a similar bump to farm earnings, it’s likely that farm earnings has a significant influence in that region’s economy.

The overall trend is that the farm earnings “bump” only appears in a few select regions and in regions that have a lot of “entirely rural” counties in them.

In addition, these “bumps” that do appear in total earnings has significantly impacted the earnings-performance of these groups/regions. For example, the entirely rural county group’s 2018 total earnings equals the value they had in 2010. This is also true for the Southwest region. In particular, EDRs 8 and 6W have total earnings equal to 2010 values.

<br>

```{r prep ag  wages and total wages index, include = FALSE, cache=TRUE}
ag.total.wages.ruca <- ag.ind.wages.county %>%
  group_by(periodyear, Dem_Desc) %>%
  summarise(ag.wages = sum(wages[naicstitle != "Total, All Industries"], na.rm = TRUE),
            total.wages = sum(wages[naicstitle == "Total, All Industries"], na.rm = TRUE)) %>%
  ungroup() %>%
  group_by(Dem_Desc) %>%
  mutate(ag.wages.index = ag.wages / ag.wages[periodyear == 2001],
         total.wages.index = total.wages / total.wages[periodyear == 2001]) %>%
  ungroup() %>%
  mutate(Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

ag.total.wages.pr <- ag.ind.wages.pr %>%
  group_by(periodyear, planning.region) %>%
  summarise(ag.wages = sum(wages[naicstitle != "Total, All Industries"], na.rm = TRUE),
            total.wages = sum(wages[naicstitle == "Total, All Industries"], na.rm = TRUE)) %>%
  ungroup() %>%
  group_by(planning.region) %>%
  mutate(ag.wages.index = ag.wages / ag.wages[periodyear == 2001],
         total.wages.index = total.wages / total.wages[periodyear == 2001]) %>%
  ungroup() %>%
  mutate(planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"))

ag.total.wages.edr <- ag.ind.wages.edr %>%
  group_by(periodyear, edr) %>%
  summarise(ag.wages = sum(wages[naicstitle != "Total, All Industries"], na.rm = TRUE),
            total.wages = sum(wages[naicstitle == "Total, All Industries"], na.rm = TRUE)) %>%
  ungroup() %>%
  group_by(edr) %>%
  mutate(ag.wages.index = ag.wages / ag.wages[periodyear == 2001],
         total.wages.index = total.wages / total.wages[periodyear == 2001]) %>%
  ungroup() %>%
  mutate(edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

```

### RUCA

<br>

There are two distinct patterns here that are dependent on the type of RUCA group. The entirely rural group and, to lesser extent, the town/rural county group saw some rather big decreases in ag wages in the early 2000's, a recovery betweene 2006 and 2015, and a subsequent downturn since then. The other county groups, entirely urban, urban/town/rural mix, and the town/rural mix have experienced significant growth in ag-related wages. In fact, total wages earned in ag-related industries has grown by nearly 200% while total wages has grown by about 175%. 

One other interesting element concerning the ag-related wages is that most of the growth occurred during the Great Recession. 

One last interesting nuance is the slight downturn in ag-related wages earned over the last few years in the entirely urban county group.

There isn't a lot of evidence that ag-related wages has impacted total wages in any of the RUCA groups. As expected, total wages plateaud during the Great Recession and has experienced growth since 2012. The entirely rural RUCA group does seem to be lagging behind a bit in total wage growth compared to the other regions. 

<br>

```{r charts ag wages and total wages index ruca, fig.width=8}
ag.wages.index.ruca.plot <- ggplot(ag.total.wages.ruca, aes(periodyear, ag.wages.index, color = Dem_Desc)) +
  geom_smooth(se = FALSE, size = 2) +
  geom_hline(yintercept = 1, color = "black") +
  geom_point_interactive(size = 3, aes(data_id = ag.wages.index, tooltip = paste(Dem_Desc, "\nYear: ", periodyear, "\nTotal ag-related wages: ", dollar(ag.wages), "\nAg wages as a % of 2001 levels: ", percent(ag.wages.index), sep = ""))) +
  labs(x="", y = "", color="", title = "Ag-related wages indexed to 2001 levels")+
  scale_y_continuous(labels=scales::percent)+
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_manual(values= color.ruca,
                     guide = guide_legend(ncol = 4)) +
  theme(text = element_text(size = 10),
        plot.margin = unit(c(0,0,0,0), "cm"))


total.wages.index.ruca.plot <- ggplot(ag.total.wages.ruca, aes(periodyear, total.wages.index, color = Dem_Desc)) +
  geom_smooth(se = FALSE, size = 2) +
  geom_hline(yintercept = 1, color = "black") +
  geom_point_interactive(size = 3, aes(data_id = total.wages.index, tooltip = paste(Dem_Desc, "\nYear: ", periodyear, "\nTotal wages: ", dollar(total.wages), "\nTotal wages as a % of 2001 levels: ", percent(total.wages.index), sep = ""))) +
  labs(x="", y = "", color="", title = "Total wages indexed to 2001 levels")+
  scale_y_continuous(labels=scales::percent)+
  coord_cartesian(ylim = seq(.5, 2, .5)) +
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_manual(values= color.ruca) +
  theme(legend.position = "none",
        text = element_text(size = 10),
        plot.margin = unit(c(0,0,0,0), "cm"))

ag.wages.index.ruca.legend <- get_legend(ag.wages.index.ruca.plot + theme(legend.direction = "horizontal", legend.justification = "right", legend.box.just = "right"))

ag.wages.index.ruca.legend.bottom <- plot_grid(NULL, ag.wages.index.ruca.legend, NULL)


girafe( ggobj = plot_grid(ag.wages.index.ruca.plot + theme(legend.position = "none"), total.wages.index.ruca.plot+theme(legend.position = "none"),  ag.wages.index.ruca.legend.bottom, nrow = 2, rel_heights = c(2,1)), width_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```

### Planning Region

<br>

```{r charts ag wages and total wages index pr, fig.width=8}
ag.wages.index.pr.plot <- ggplot(ag.total.wages.pr, aes(periodyear, ag.wages.index, color = planning.region)) +
  geom_smooth(se = FALSE, size = 2) +
  geom_hline(yintercept = 1, color = "black") +
  geom_point_interactive(size = 3, aes(data_id = ag.wages.index, tooltip = paste(planning.region, "\nYear: ", periodyear, "\nTotal ag-related wages: ", dollar(ag.wages), "\nAg wages as a % of 2001 levels: ", percent(ag.wages.index), sep = ""))) +
  labs(x="", y = "", color="", title = "Ag-related wages indexed to 2001 levels")+
  scale_y_continuous(labels=scales::percent)+
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_manual(values= color.pr,
                     guide = guide_legend(ncol = 4)) +
  theme(text = element_text(size = 10),
        plot.margin = unit(c(0,0,0,0), "cm"))


total.wages.index.pr.plot <- ggplot(ag.total.wages.pr, aes(periodyear, total.wages.index, color = planning.region)) +
  geom_smooth(se = FALSE, size = 2) +
  geom_hline(yintercept = 1, color = "black") +
  geom_point_interactive(size = 3, aes(data_id = total.wages.index, tooltip = paste(planning.region, "\nYear: ", periodyear, "\nTotal wages: ", dollar(total.wages), "\nTotal wages as a % of 2001 levels: ", percent(total.wages.index), sep = ""))) +
  labs(x="", y = "", color="", title = "Total wages indexed to 2001 levels")+
  scale_y_continuous(labels=scales::percent)+
  coord_cartesian(ylim = seq(.5, 2, .5)) +
  scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_manual(values= color.pr) +
  theme(legend.position = "none",
        text = element_text(size = 10),
        plot.margin = unit(c(0,0,0,0), "cm"))

ag.wages.index.pr.legend <- get_legend(ag.wages.index.pr.plot + theme(legend.direction = "horizontal", legend.justification = "right", legend.box.just = "right"))

ag.wages.index.pr.legend.bottom <- plot_grid(NULL, ag.wages.index.pr.legend, NULL)


girafe( ggobj = plot_grid(ag.wages.index.pr.plot + theme(legend.position = "none"), total.wages.index.pr.plot+theme(legend.position = "none"),  ag.wages.index.pr.legend.bottom, nrow = 2, rel_heights = c(2,1)), width_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```


