---
title: "Workspace"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
runtime: shiny
resource_files:
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.cpg
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.dbf
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.prj
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbn
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbx
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shp.xml
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shx
---

<link rel="stylesheet" href="styles.css" type="text/css">


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
```

```{r loading jon docs and shapefiles, cache=TRUE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", size = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.title = element_blank(),
        text = element_text(size = 15),
        plot.caption = element_text(hjust = 0.5, face = "italic"),
        legend.text = element_text(margin = margin(l = 2, r = 5)))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", size = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        text = element_text(size = 15),
        plot.caption = element_text(hjust = 0.5, face = "italic"),
        legend.text = element_text(margin = margin(l = 2, r = 5)))

theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        plot.title = element_text(hjust = 0.5),
        legend.title = element_blank(),
        text = element_text(size = 15),
        plot.caption = element_text(hjust = 0.5, face = "italic"),
        legend.text = element_text(margin = margin(l = 2, r = 5)))

counties.regions <- read_csv("Data/Join docs/county_regions.csv") %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc))

regions = counties.regions %>%
  select(5,6) %>%
  distinct(edr, .keep_all = TRUE)

edr.pr <- counties.regions %>%
  distinct(edr, .keep_all = TRUE) %>%
  select(5,6) %>%
  mutate(edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""))

color.ruca <- c("Entirely rural" = "#5CA81F", "Town/rural mix" = "#C7EF99", "Urban/town/rural mix" = "#d8b365", "Entirely urban" = "#a6611a", "Minnesota" = "black")

color.pr <- c("Northwest" = "#810f7c","Northeast" = "#fe9929", "Central" = "#076324", "Seven County Mpls-St Paul" = "#d8b365", "Southwest" = "#1f78b4", "Southeast" = "#d7301f", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.counties <- scale_color_brewer(palette = "Dark2",
                       guide = guide_legend(ncol = 3))

mn_counties <- st_read("Data/Shapefiles/county shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE)
```

```{r master objects, include=FALSE, cache=TRUE}
master.qcew.pr <- read_csv("Data/Lazarus/lazarus-qcew-pr.csv") %>%
  rename(planning.region = areaname) %>%
  mutate(planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = str_replace(planning.region, ", MN", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"))

master.qcew.edr <- read_csv("Data/Lazarus/lazarus-qcew-edr.csv") %>%
  mutate(areaname = str_replace(areaname, "  ", " ")) %>%
  rename("edr" = "areaname") %>%
  left_join(regions, by = "edr") %>%
  mutate(planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

master.qcew.county <- read_csv("Data/Lazarus/lazarus-qcew-county.csv") %>%
  mutate(areaname = str_replace(areaname, " County", ""),
         areaname = str_replace(areaname, "Saint Louis", "St. Louis")) %>%
  left_join(counties.regions[,c(1,2)], by = c("areaname" = "Name"))
```

```{r prep combined ag industries as pct of total, include=FALSE, cache=TRUE}

emp.pr <- master.qcew.pr %>%
  select(1,2,5,28) %>%
  group_by(periodyear, planning.region) %>%
  summarise(ag.emp = sum(empYear[naicstitle != "Total, All Industries"], na.rm = TRUE),
            total.emp = sum(empYear[naicstitle == "Total, All Industries"], na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(ag.pct.total = ag.emp / total.emp)

emp.edr <- master.qcew.edr %>%
  select(1,2,5,28, 83) %>%
  group_by(periodyear, edr, planning.region) %>%
  summarise(ag.emp = sum(empYear[naicstitle != "Total, All Industries"], na.rm = TRUE),
            total.emp = sum(empYear[naicstitle == "Total, All Industries"], na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(ag.pct.total = ag.emp / total.emp)

emp.county <- master.qcew.county %>%
  select(1,2,5,28,83) %>%
  group_by(periodyear, areaname, countyfp) %>%
  summarise(ag.emp = sum(empYear[naicstitle != "Total, All Industries"], na.rm = TRUE),
            total.emp = sum(empYear[naicstitle == "Total, All Industries"], na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(ag.pct.total = ag.emp / total.emp,
         ag.pct.total.bins = cut(ag.pct.total,
                                 breaks = c(-1, .01, .02, .03, .04, 1),
                                 labels = c("0% to 1%", "1% to 2%", "2% to 3%", "3% to 4%", "More than 4%"))) %>%
  left_join(mn_counties[,c(4,7)], by = c("countyfp" = "FIPS_CODE")) 
```

```{r prep filtered counties, include=FALSE, cache=TRUE}
filt.counties <- emp.county %>%
  filter(ag.pct.total > .0299999) %>%
  distinct(areaname, .keep_all = TRUE)

filt.trends.counties <- emp.county %>%
  right_join(filt.counties[,c(3)], by = "countyfp") %>%
  group_by(areaname, countyfp) %>%
  mutate(ag.emp.change = ag.emp / ag.emp[periodyear==2001]) %>%
  left_join(counties.regions[,c(1,4,5,6)], by = "countyfp")
```

<br>

## Specific ag industries driving change {.tabset}

The next question to ask is what industries identified in the Lazarus report are driving the increases and decreases in those ag-related employment sectors. Below are the 16 counties identified above experiencing somewhat clear trends of decreasing and increasing employment in the industries identified in the Lazarus study and the employment trends in each of those industries.

Due to such low numbers of employment, the data is a bit spotty and only a few industries actually show up at the county level. This, in itself, is indicative of the lack of impact these industries really have on overall employment in Minnesota. 

Overall, there are two main industries that seem to indicate whether a county is increasing or decreasing in employment in these ag-related industries as a whole;

1. farm product raw material merchant wholesalers, and
2. farm supplies merchant wholesalers.

If these industries were increasing in the county, than it was likely that the county was experiencing an increase in the ag-related industries as a whole, and vice-versa.

<br>

```{r prep industry trends for impacted counties, include=FALSE, cache = TRUE}
impacted.counties <- master.qcew.county %>%
  filter(areaname %in% c("Grant", "Stevens", "McLeod", "Swift", "Chippewa", "Pipestone", "Rock", "Martin", "Faribault", "Marshall", "Wilkin", "Meeker", "Renville", "Lac qui Parle", "Redwood", "Watonwan"),
         naicstitle != "Total, All Industries") %>%
  select(1,2,5,28,83) %>%
  group_by(areaname, countyfp) %>%
  complete(periodyear, naicstitle, fill = list(c(empYear = 0))) %>%
  mutate(empYear = ifelse(is.na(empYear), 0, empYear),
         inc.dec = ifelse(areaname %in% c("Grant", "Stevens", "McLeod", "Swift", "Chippewa", "Pipestone", "Rock", "Martin", "Faribault"), "Decreasing", "Increasing"),
         data_id = seq(n())) %>%
  left_join(counties.regions[,c(1,5,6)], by = "countyfp")

```

### Counties with increasing employment in Lazarus industries{.tabset}

In counties where there is increasing employment in industries identified in the Lazarus study, there seems to be two industries within the retail and trade sector driving the increases;

1. Farm product raw material merchant wholesales, and
2. Farm supplies merchant wholesalers.

<br>

```{r charts lazarus industry trends increasing counties, fig.height=9}

impacted.counties.plot <- ggplot(filter(impacted.counties, inc.dec == "Increasing"), aes(periodyear, empYear, color = naicstitle)) +
  facet_wrap(~areaname, ncol = 2) +
  geom_point_interactive(size = 2, aes(data_id = data_id, tooltip = paste(areaname, "Year: ", periodyear, "\nNAICS title: ", naicstitle, "\nEmployment: ", comma(empYear), sep = ""))) +
  labs(x="", y = "", color="", title = "Employment by industry and county")+
  scale_y_continuous(labels=scales::comma)+
    scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_discrete(guide = guide_legend(ncol = 2)) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        legend.text = element_text(size = 8),
        text = element_text(size = 10))

girafe(ggobj = impacted.counties.plot, height_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```


### Counties with decreasing employment in Lazarus industries

The overall industry driver of declines these counties are;

* farm and garden machinery and equipment merchant wholesalers,
* farm product raw material merchant wholesalers
* farm supplies merchant wholesalers.

Below is a list of each of the counties and the general trends of their industries.

Chippewa: Decreases in farm and garden machinery and equipment merchant wholesalers and farm supplies merchant wholesalers.

Faribault: Declines in Farm product raw material merchant wholesalers, and farm and garden machinery and equipment merchant wholesalers.

Grant: Declines in farm product raw material merchant wholesalers and the ups and downs of farm supplies merchant wholesalers.

Martin: Significant appearances and disappearances of farm supplies merchant wholesalers and farm product raw material merchant wholesalers.

McLeod: Disappearance of farm and garden machinery and equipment merchant wholesalers, and ag implement manufacturing. But seeing increases in farm supplies merchant wholesalers and commercial and institutional building construction.

Pipestone: Decline in farm product raw materials merchant wholesalers and disappearance of farm and garden machinery and equipment merchant wholesalers.

Rock: Disappearance of farm and garden machinery and equipment merchant wholesalers and farm supplies merchant wholesaler.

Stevens: Decline in farm product raw materials merchant wholesalers with some ups and downs in the farm and garden machinery and equipment merchant wholesalers. Increases in farm supplies merchant wholesalers.

Swift: Decreases in farm product raw material merchant wholesalers.


<br>

```{r charts lazarus industry trends decreasing counties, fig.height=9}

impacted.counties.decreasing.plot <- ggplot(filter(impacted.counties, inc.dec == "Decreasing"), aes(periodyear, empYear, color = naicstitle)) +
  facet_wrap(~areaname, ncol = 2) +
  geom_point_interactive(size = 2, aes(data_id = data_id, tooltip = paste(areaname, "Year: ", periodyear, "\nNAICS title: ", naicstitle, "\nEmployment: ", comma(empYear), sep = ""))) +
  labs(x="", y = "", color="", title = "Employment by industry and county")+
  scale_y_continuous(labels=scales::comma)+
    scale_x_continuous(breaks = seq(1900, 2050, 2)) +
  theme_line+
  scale_color_discrete(guide = guide_legend(ncol = 2)) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        legend.text = element_text(size = 8),
        text = element_text(size = 10))

girafe(ggobj = impacted.counties.decreasing.plot, height_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE) ,
                 opts_sizing(rescale = FALSE))

```
