---
title: "Workspace"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
runtime: shiny
resource_files:
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.cpg
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.dbf
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.prj
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbn
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbx
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shp.xml
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
```

```{r loading jon docs and shapefiles, cache=TRUE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", size = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.title = element_blank(),
        text = element_text(size = 15),
        plot.caption = element_text(hjust = 0.5, face = "italic"),
        legend.text = element_text(margin = margin(l = 2, r = 5)))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", size = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        text = element_text(size = 15),
        plot.caption = element_text(hjust = 0.5, face = "italic"),
        legend.text = element_text(margin = margin(l = 2, r = 5)))

theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        plot.title = element_text(hjust = 0.5),
        legend.title = element_blank(),
        text = element_text(size = 15),
        plot.caption = element_text(hjust = 0.5, face = "italic"),
        legend.text = element_text(margin = margin(l = 2, r = 5)))

counties.regions <- read_csv("Data/Join docs/county_regions.csv") %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc))

edr.pr <- counties.regions %>%
  distinct(edr, .keep_all = TRUE) %>%
  select(5,6) %>%
  mutate(edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""))

color.ruca <- c("Entirely rural" = "#5CA81F", "Town/rural mix" = "#C7EF99", "Urban/town/rural mix" = "#d8b365", "Entirely urban" = "#a6611a", "Minnesota" = "black")

color.pr <- c("Northwest" = "#810f7c","Northeast" = "#fe9929", "Central" = "#076324", "Seven County Mpls-St Paul" = "#d8b365", "Southwest" = "#1f78b4", "Southeast" = "#d7301f", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.counties <- scale_color_brewer(palette = "Dark2",
                       guide = guide_legend(ncol = 3))

mn_counties <- st_read("Data/Shapefiles/county shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE)
```

```{r master land value and net income objects, include=FALSE, cache=TRUE}
master.netinc.county <- read_csv("Data/Income and expenses/Master-farm-income-expenses.csv") %>%
  mutate(Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = str_replace(edr, "  ", " "),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota")) %>%
  left_join(counties.regions[,c(1,2)], by = "countyfp") %>%
  filter(Name != "Minnesota")

master.lv.county <- read_csv("Data/Land values/Master-ag-land-values.csv") %>%
  mutate(Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = str_replace(edr, "  ", " "),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota")) %>%
  rename(year = Year) %>%
  filter(County != "Minnesota") %>%
  rename(Name = County)

```

```{r prep net income and land values index, include=FALSE, cache=TRUE}
net.income.acre.ruca <- master.netinc.county %>%
  group_by(Dem_Desc, year) %>%
  summarise(net.income = sum(net.income, na.rm = TRUE),
            total.acres = sum(Total.acres, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(net.income.acre = net.income / total.acres) %>%
  group_by(Dem_Desc) %>%
  mutate(net.income.acre.index = net.income.acre / net.income.acre[year == 2001]) %>%
  ungroup()

land.value.acre.ruca <- master.lv.county %>%
  group_by(Dem_Desc, year) %>%
  summarise(total.acres = sum(Total.acres, na.rm = TRUE),
            total.value = sum(Total.estimated.value, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(land.value.acre = total.value / total.acres) %>%
  group_by(Dem_Desc) %>%
  mutate(land.value.acre.index = land.value.acre / land.value.acre[year == 2001]) %>%
  ungroup()

net.income.acre.pr <- master.netinc.county %>%
  group_by(planning.region, year) %>%
  summarise(net.income = sum(net.income, na.rm = TRUE),
            total.acres = sum(Total.acres, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(net.income.acre = net.income / total.acres) %>%
  group_by(planning.region) %>%
  mutate(net.income.acre.index = net.income.acre / net.income.acre[year == 2001]) %>%
  ungroup()

land.value.acre.pr <- master.lv.county %>%
  group_by(planning.region, year) %>%
  summarise(total.acres = sum(Total.acres, na.rm = TRUE),
            total.value = sum(Total.estimated.value, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(land.value.acre = total.value / total.acres) %>%
  group_by(planning.region) %>%
  mutate(land.value.acre.index = land.value.acre / land.value.acre[year == 2001]) %>%
  ungroup()

net.income.acre.edr <- master.netinc.county %>%
  group_by(edr, planning.region, year) %>%
  summarise(net.income = sum(net.income, na.rm = TRUE),
            total.acres = sum(Total.acres, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(net.income.acre = net.income / total.acres) %>%
  group_by(edr, planning.region) %>%
  mutate(net.income.acre.index = net.income.acre / net.income.acre[year == 2001]) %>%
  ungroup()

land.value.acre.edr <- master.lv.county %>%
  group_by(edr, planning.region, year) %>%
  summarise(total.acres = sum(Total.acres, na.rm = TRUE),
            total.value = sum(Total.estimated.value, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(land.value.acre = total.value / total.acres) %>%
  group_by(planning.region) %>%
  mutate(land.value.acre.index = land.value.acre / land.value.acre[year == 2001]) %>%
  ungroup()

```

```{r prep year of peak net income, include=FALSE, cache=TRUE}
net.inc.peak.year.ruca <- net.income.acre.ruca %>%
  group_by(Dem_Desc) %>%
  top_n(n = 1, wt = net.income.acre) %>%
  ungroup() %>%
  rename(peak.year = 2,
         net.income.peak = 3,
         total.acre.peak = 4,
         net.income.acre.peak = 5)

land.value.peak.year.ruca <- land.value.acre.ruca %>%
  right_join(net.inc.peak.year.ruca[,c(1,2)], by = c("Dem_Desc", "year" = "peak.year")) %>%
  rename(peak.year = 2,
         total.acres.peak = 3,
         total.value.peak = 4,
         land.value.acre.peak = 5)


net.inc.change.peak.year.ruca <- net.income.acre.ruca %>%
  filter(year == 2018) %>%
  left_join(net.inc.peak.year.ruca[,c(1,2,3,4,5)], by = "Dem_Desc") %>%
  mutate(pct.change.net.income.acre = (net.income.acre - net.income.acre.peak) / net.income.acre.peak)

land.value.change.peak.year.ruca <- land.value.acre.ruca %>%
  filter(year == 2018) %>%
  left_join(land.value.peak.year.ruca[,c(1,2,3,4,5)], by = "Dem_Desc") %>%
  mutate(pct.change.land.value.acre = (land.value.acre - land.value.acre.peak) / land.value.acre.peak)

net.inc.land.value.change.peak.year.ruca <- net.inc.change.peak.year.ruca %>%
  left_join(land.value.change.peak.year.ruca[,c(1,4,5,9,10,11)], by = "Dem_Desc") %>%
  gather(key = "income.land", value = "pct.change", 11, 16) %>%
  mutate(income.land = ifelse(income.land == "pct.change.net.income.acre", "Change in net farm income per acre", "Change in land value per farm acre"))


net.inc.peak.year.pr <- net.income.acre.pr %>%
  group_by(planning.region) %>%
  top_n(n = 1, wt = net.income.acre) %>%
  ungroup() %>%
  rename(peak.year = 2,
         net.income.peak = 3,
         total.acre.peak = 4,
         net.income.acre.peak = 5)

land.value.peak.year.pr <- land.value.acre.pr %>%
  right_join(net.inc.peak.year.pr[,c(1,2)], by = c("planning.region", "year" = "peak.year")) %>%
  rename(peak.year = 2,
         total.acres.peak = 3,
         total.value.peak = 4,
         land.value.acre.peak = 5)


net.inc.change.peak.year.pr <- net.income.acre.pr %>%
  filter(year == 2018) %>%
  left_join(net.inc.peak.year.pr[,c(1,2,3,4,5)], by = "planning.region") %>%
  mutate(pct.change.net.income.acre = (net.income.acre - net.income.acre.peak) / net.income.acre.peak)

land.value.change.peak.year.pr<- land.value.acre.pr %>%
  filter(year == 2018) %>%
  left_join(land.value.peak.year.pr[,c(1,2,3,4,5)], by = "planning.region") %>%
  mutate(pct.change.land.value.acre = (land.value.acre - land.value.acre.peak) / land.value.acre.peak)

net.inc.land.value.change.peak.year.pr <- net.inc.change.peak.year.pr %>%
  left_join(land.value.change.peak.year.pr[,c(1,4,5,9,10,11)], by = "planning.region") %>%
  gather(key = "income.land", value = "pct.change", 11, 16) %>%
  mutate(income.land = ifelse(income.land == "pct.change.net.income.acre", "Change in net farm income per acre", "Change in land value per farm acre"),
         planning.region = str_replace(planning.region, "Seven County Mpls-St Paul", "Seven County\nMpls-St Paul"))



net.inc.peak.year.edr <- net.income.acre.edr %>%
  group_by(edr, planning.region) %>%
  top_n(n = 1, wt = net.income.acre) %>%
  ungroup() %>%
  rename(peak.year = 3,
         net.income.peak = 4,
         total.acre.peak = 5,
         net.income.acre.peak = 6)

land.value.peak.year.edr <- land.value.acre.edr %>%
  right_join(net.inc.peak.year.edr[,c(1,2,3)], by = c("edr", "planning.region", "year" = "peak.year")) %>%
  rename(peak.year = 3,
         total.acres.peak = 4,
         total.value.peak = 5,
         land.value.acre.peak = 6)

net.inc.change.peak.year.edr <- net.income.acre.edr %>%
  filter(year == 2018) %>%
  left_join(net.inc.peak.year.edr[,c(1,2,3,4,5,6)], by = c("planning.region", "edr")) %>%
  mutate(pct.change.net.income.acre = (net.income.acre - net.income.acre.peak) / net.income.acre.peak)

land.value.change.peak.year.edr<- land.value.acre.edr %>%
  filter(year == 2018) %>%
  left_join(land.value.peak.year.edr[,c(1,2,3,4,5,6)], by = c("planning.region", "edr")) %>%
  mutate(pct.change.land.value.acre = (land.value.acre - land.value.acre.peak) / land.value.acre.peak)

net.inc.land.value.change.peak.year.edr <- net.inc.change.peak.year.edr %>%
  left_join(land.value.change.peak.year.edr[,c(1,2,5, 6,10,11,12)], by = c("edr", "planning.region")) %>%
  gather(key = "income.land", value = "pct.change", 12, 17) %>%
  mutate(income.land = ifelse(income.land == "pct.change.net.income.acre", "Change in net farm income per acre", "Change in land value per farm acre"))

```

<br>

## Counties - land values most negatively impacted by farm incomes

As the previous charts have shown, land values have remained stubbornly high despite the massive change in net farm incomes. Most of the regions in the state actually have higher land values in 2018 than when net farm incomes peaked. This could be due to a slower response by land values, and they increased for a few years after net farm incomes peaked, but have been declining over the last few years but just happen to still be above values when incomes peaked. But our concern here is identifying counties where land values have decreased enough where they are below values when incomes peaked.

The previous charts have provided us charatertics to look for in counties that are likely being negatively impacted by decreasing land values;

* 2018 agricultural land values that are below their values,
* likely in our most rural counties,
* and located in EDR 6W, EDR 8, and EDR 7E.

Summary here

```{r prep land value map, include = FALSE, cache=TRUE}

farm.inc.peak.county <- master.netinc.county %>%
  mutate(net.income.acre = net.income / Total.acres) %>%
  group_by(countyfp, Name) %>%
  top_n(n = 1, wt = net.income.acre) %>% 
  arrange(desc(year)) %>%
  distinct(Name, .keep_all = TRUE) %>%
  ungroup() %>%
  rename(peak.year = 1,
         total.acres.peak = 7,
         net.income.acre.peak = 13)

names(farm.inc.peak.county)
```


<br>


